---
title: "Project 1"
author: "Björn Ahl, Leon Krasniqi, Nora Granström"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
execute: 
  echo: false
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(car)
library(rstatix)
library(GGally)
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "/Users/noragranstrom/Documents/GitHub/MASM22/Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

# Part 1. Particle emissions and fuel consumption

## 1(a) Choice of transformations

We compare residual plots for three candidate models to motivate the log of the response and to decide between x = Fuel and x = log(Fuel).

```{r}
#| label: fig-1a-resid
#| fig-cap: "Residuals vs fitted for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

library(broom)
library(patchwork)

m_raw    <- lm(PM10      ~ Fuel,       data = df)
m_logy   <- lm(log(PM10) ~ Fuel,       data = df)
m_loglog <- lm(log(PM10) ~ log(Fuel),  data = df)

plot_resid <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = .6) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x, colour = "steelblue") +
    labs(title = title, x = "fitted", y = "residual")
}

plot_resid(m_raw,    "PM10 ~ Fuel") +
  plot_resid(m_logy,   "log(PM10) ~ Fuel") +
  plot_resid(m_loglog, "log(PM10) ~ log(Fuel)")
```

**Interpretation:**

- **Linear relationship.** In this case we have PM10 = x₀ + x₁·Fuel. From the graph we observe a stronger tendency toward the left, meaning observations cluster at lower fuel values while the spread of PM10 widens with Fuel. This suggests a purely linear model is not sufficient.
- **Logarithmic response.** ln(PM10) = x₀ + x₁·Fuel is equivalent to PM10 = a·b\^Fuel. The residual plot is less skewed, but equal additive changes in Fuel are treated the same across the scale, which is unrealistic for municipalities of different sizes.
- **Log-log relationship.** ln(PM10) = x₀ + x₁·ln(Fuel) implies PM10 = a·Fuel\^x₁. A relative change in Fuel implies a relative change in PM10, and the residuals are more evenly spread. This makes the log-log transformation the most plausible choice.

```{r}
#| label: fig-1a-qq
#| fig-cap: "QQ-plots of residuals for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

plot_qq <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(sample = .resid)) +
    stat_qq(alpha = .6) +
    stat_qq_line(colour = "steelblue") +
    labs(title = title, x = "theoretical quantiles", y = "sample quantiles")
}
plot_qq(m_raw,    "PM10 ~ Fuel") +
    plot_qq(m_logy,   "log(PM10) ~ Fuel") +
    plot_qq(m_loglog, "log(PM10) ~ log(Fuel)")
```

From the QQ-plots we can clearly see that the log–log transformation gives the best fit, as its residuals follow the reference line most closely and are therefore the most normally distributed of the three.

## 1(b) Confidence interval, prediction interval and $\beta$-estimates

In this part-question, we obtain the $\beta$-estimates and their 95% confidence intervals when `x = ln(Fuel)`.

```{r}
#| label: fig-1b-model

mymodel <- lm(PM10 ~ log(Fuel), data = df)
mymodel_summary <- summary(mymodel)
mymodel$coefficients
confint(mymodel)

another_model <- lm(log(PM10) ~ log(Fuel), data = df)

new_data <- data.frame(Fuel = seq(min(df$Fuel), max(df$Fuel), length.out = 100))

pred_int <- predict(another_model, newdata = new_data, interval = "prediction")
conf_int <- predict(another_model, newdata = new_data, interval = "confidence")

plot_df <- data.frame(
  Fuel = new_data$Fuel,
  fit = conf_int[, "fit"],
  ci_lower = conf_int[, "lwr"],
  ci_upper = conf_int[, "upr"],
  pi_lower = pred_int[, "lwr"],
  pi_upper = pred_int[, "upr"]
)

# Plot on log scale
p1 <- ggplot() +
  geom_point(data = df, aes(x = log(Fuel), y = log(PM10))) +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df, aes(x = log(Fuel), y = fit), color = "red") +
  labs(x = "ln(Fuel)", y = "ln(PM10)",
       title = "Log scale: ln(PM10) ~ ln(Fuel)",
       subtitle = "red = fit, dark band = 95% CI, light band = 95% PI") +
  theme_minimal()

# Transforming back to original scale
plot_df_e <- data.frame(
  Fuel = new_data$Fuel,
  fit = exp(conf_int[, "fit"]),
  ci_lower = exp(conf_int[, "lwr"]),
  ci_upper = exp(conf_int[, "upr"]),
  pi_lower = exp(pred_int[, "lwr"]),
  pi_upper = exp(pred_int[, "upr"])
)

p2 <- ggplot() +
  geom_point(data = df, aes(x = Fuel, y = PM10)) +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df_e, aes(x = Fuel, y = fit), color = "red") +
  labs(x = "Fuel", y = "PM10",
       title = "Original scale: PM10 vs Fuel (back-transformed)",
       subtitle = "red = fit, dark band = 95% CI, light band = 95% PI") +
  theme_minimal()

# Show both side by side so the log-scale and back-transformed versions are easy to compare
p1 + p2 + plot_annotation(title = "Model 1(b): ln(PM10) = beta0 + beta1 * ln(Fuel)")

# Back-transformed relationship: PM10 = exp(beta0) * Fuel^beta1 = a * Fuel^beta1
intercept <- cbind(
  expbeta = exp(another_model$coefficients["(Intercept)"]),
  exp(confint(another_model, parm = "(Intercept)"))
) |>
  round(digits = 4)

slope <- cbind(
  expbeta = exp(another_model$coefficients["log(Fuel)" ]),
  exp(confint(another_model, parm = "log(Fuel)"))
) |>
  round(digits = 4)

intercept   # a  = exp(beta0) with 95% CI
slope       # exp(beta1) with 95% CI (the elasticity factor per unit change in log(Fuel))
```

**Back-transformation.** Exponentiating both sides of $\ln(\text{PM10}) = \beta_0 + \beta_1\ln(\text{Fuel})$ gives
$$\text{PM10} = e^{\beta_0}\cdot\text{Fuel}^{\beta_1} = a\cdot\text{Fuel}^{\beta_1}, \qquad a = e^{\beta_0}.$$

The plug-in estimates are $\widehat{a} = e^{\widehat{\beta}_0}$ (with 95 % CI $[\,e^{L_0},\,e^{U_0}\,]$) and $\widehat{\beta}_1$ with its usual CI — these are printed as `intercept` and `slope` above. Note that the CI/PI ribbons in the PM10 vs Fuel plot are the exponentials of the ribbons in the ln–ln plot, so they are valid intervals on the original scale but are no longer symmetric about the fit.

**Model-fit comments.** On the ln–ln scale (p1) the points scatter reasonably evenly around the line and the ribbons hug the cloud well — the assumption of constant variance looks plausible there. On the back-transformed plot (p2) the prediction band fans out dramatically at high Fuel, reflecting that constant variance on the log scale is multiplicative variance on the original scale; a handful of high-Fuel municipalities sit well outside the confidence band, suggesting leverage/outlier influence. In the residual plot from 1(a), the log–log residuals are the most evenly spread of the three candidates, but there is still some funneling and a few large negative residuals, signalling that `log(Fuel)` alone does not capture all the structure — more x-variables will be added in Part 2.

## 1(c) Effect of changing fuel consumption

Model 1(b) is the log–log model $$\ln(\text{PM10}) = \beta_0 + \beta_1 \ln(\text{Fuel}) \;\Longleftrightarrow\; \text{PM10} = a \cdot \text{Fuel}^{\beta_1}, \quad a = e^{\beta_0}.$$

Here $\beta_1$ is an elasticity: if Fuel is multiplied by a factor $k$, PM10 is multiplied by $k^{\beta_1}$.

**Change rate for a 10 % decrease in Fuel.** Let $k = 0.9$, so the new Fuel is $0.9\,\text{Fuel}_{\text{old}}$. Then $$\frac{\text{PM10}_{\text{new}}}{\text{PM10}_{\text{old}}} = \left(\frac{0.9\,\text{Fuel}_{\text{old}}}{\text{Fuel}_{\text{old}}}\right)^{\beta_1} = 0.9^{\beta_1},$$ giving the point estimate $\widehat{r} = 0.9^{\widehat{\beta}_1}$ and percent change $(0.9^{\widehat{\beta}_1}-1)\cdot 100\%$.

With the 95 % CI for $\beta_1$ given by $[L,\,U] = \widehat{\beta}_1 \pm t_{0.975,\,n-2}\cdot \text{SE}(\widehat{\beta}_1)$, the map $\beta_1 \mapsto 0.9^{\beta_1}$ is **decreasing** (since $0.9 < 1$), so $$\text{CI}_{95\%}(r) = \bigl[\,0.9^{U},\;0.9^{L}\,\bigr].$$

**Fuel reduction required to halve PM10.** We solve $k^{\beta_1} = 0.5$ for $k$: $$k = 0.5^{1/\beta_1}, \qquad \text{required reduction} = (1 - 0.5^{1/\beta_1})\cdot 100\%.$$

The map $\beta_1 \mapsto 0.5^{1/\beta_1}$ is **increasing** for $\beta_1 > 0$, so $$\text{CI}_{95\%}(k) = \bigl[\,0.5^{1/L},\;0.5^{1/U}\,\bigr],$$ and the CI for the reduction is $\bigl[\,1 - 0.5^{1/U},\;1 - 0.5^{1/L}\,\bigr]\cdot 100\%.$

```{r}
#| label: q1c
#| message: false
#| warning: false

b1  <- coef(another_model)["log(Fuel)"]
ci  <- confint(another_model, parm = "log(Fuel)")
L   <- ci[1]; U <- ci[2]

# (i) 10 % decrease in Fuel
r_hat <- 0.9^b1
r_ci  <- c(0.9^U, 0.9^L)
pct_change <- (c(r_hat, r_ci) - 1) * 100

# (ii) fuel reduction to halve PM10
k_hat <- 0.5^(1/b1)
k_ci  <- c(0.5^(1/L), 0.5^(1/U))
reduction <- (1 - c(k_hat, k_ci[2], k_ci[1])) * 100

round(rbind(
  "PM10 change (%) from -10% Fuel" = pct_change,
  "Fuel reduction (%) to halve PM10" = reduction
), 2)
```

## 1(d) Testing of the model

In this part, we are interested to test if there is a significant relationship between `log(PM10)` and `log(Fuel)`, according to the linear regression model. We can do this by using a t-test on the null hypothesis $H_0: \beta_1 = 0$ against the alternative $H_1: \beta_1 \neq 0$.

```{r}
#| label: fig-1d-testing
#| message: false
#| warning: false

another_model_summary <- summary(another_model)
another_model_summary$coefficients
another_model$df.residual

```

The value of the test-statistic for the slope coefficient `log(Fuel)` is given by the t-value in the summary output. The p-value for the slope coefficient `log(Fuel)` (Pr(\>\|t\|)=1.828846e-31) is very small (less than 0.05), which indicates that we can reject the null hypothesis and conclude that there is a statistically significant relationship between `log(PM10)` and `log(Fuel)`. We also have t ≈ 11.8 as well as 288 degrees of freedom

## 1(e) Explaining of variability

In this section, we want to determine how much of the variability in `log(PM10)` between municaplities can be explained. For that we can look at the $R^2$, which describe the proportion of variability that is explained by the model.

```{r}

# This compute R^2 for the model with log-transformation
another_model_summary$r.squared

# Adjusted R^2 for the model with log-transformation
another_model_summary$adj.r.squared

```

The coefficient of determination $R^2$ is defined as the fraction of the variability of $\\mathbf{Y}$ that is explained by the regression model $\mathbf{\widehat{Y}}=\mathbf{X}\boldsymbol{\widehat{\beta}}$. For $R^2$, we got the value $0.3772511$, for adjusted $R^2$, we got te value $0.3750888$. For $R^2$, we would like as high as possible ($0 \leq R^2 \leq 1$).

## Part 2(a) Log-transformations of numerical x-variables

We plot log(PM10) against each of the other numerical variables (Fuel through GRP) to determine which of them should also be log-transformed.

```{r}
#| label: fig-2a
#| fig-cap: "log(PM10) against each of the numerical x-variables."
#| message: false
#| warning: false

df |> select(Kommun, Fuel:PM10) |>
  pivot_longer(cols = Fuel:GRP) |>
  ggplot(aes(x = value, y = PM10)) +
  geom_point() +
  facet_wrap(~ name, scales = "free_x") +
  scale_y_log10()
```

**What to log (per PDF rules):**

- **Fuel, Builton** — points squished left as well as long right tail $\implies$ Skewed $\implies$ Log
- **Income, GRP** — economic variables (The pdf hints that we should log this) relative change more natural $\implies$ **Log.**
- **Higheds** Not really skewed, this is bounded in percentage as well. $\implies$ **No log.**
- **Vehicles, Seniors** spread evenly across axis $\implies$ **No log.**

# 2(b) Pair-wise correlations

We now want to investigate the pair-wise correlation of the numerical x variables. We want to see which ones have a strong correlation between each other, meaning an absolute value greater than or equal to 0.7. The results are visualized in the plot and summarized in a correlation matrix. The values can clearly be read off from the matrix below.

```{r}
#| label: fig-2b
#| message: false
#| warning: false

library(tidyverse)
library(GGally)

df2 <- df |> 
  select(Kommun, Fuel:GRP)

df_long <- df2 |> 
  pivot_longer(cols = Fuel:GRP) |> 
  mutate(value = log(value))

df_wide <- df_long |> 
  pivot_wider(
    id_cols = Kommun,
    names_from = name,
    values_from = value,
    names_prefix = "log_"
  )

df_wide <- df_wide |> select(-Kommun)

ggpairs(df_wide)

cor_matrix <- cor(df_wide)

```

## 2(c) $\beta$-estimates of numerical variables and computation of VIF-values.

In this section, we fit a multiple linear regression model with log(PM10) as the response variable and all seven log-transformed numerical variables to present the $\beta$-estimates and their standard errors. We will also compute VIF-values (variance inflation factors) to determine whether the predictors should remain in the model.

```{r}

model_seven <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors), data = df)
summary(model_seven)

vif(model_seven)

vif_values <- vif(model_seven)
barplot(vif_values, main = "VIF Values for Predictors", xlab = "Predictors", ylab = "VIF", col = "steelblue")
abline(h = 5, col = "red")

model_six <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Higheds) + log(Seniors), data = df) # This is without Builton
summary(model_six)

vif(model_six)

vif_values1 <- vif(model_six)
barplot(vif_values1, main = "VIF Values for Predictors", xlab = "Predictors", ylab = "VIF", col = "steelblue")
abline(h = 5, col = "red")

```

We see that log(Builton) has the largest VIF (11.12) so we drop that one first.
When fitting the model for log-$PM_{10}$ with all seven log-transformed numerical predictors, we obtain the $\beta$-estimates and their standard errors from the summary output. These are (before excluding any predictor):

| Predictor     | Estimate | Std. Error |
|---------------|----------|------------|
| Intercept     | -6.41290 | 1.46047    |
| log(Fuel)     | 0.21885  | 0.01956    |
| log(GRP)      | 0.08475  | 0.04177    |
| log(Income)   | 0.23496  | 0.24081    |
| log(Vehicles) | 0.48895  | 0.13271    |
| log(Builton)  | 0.12031  | 0.05084    |
| log(Higheds)  | -0.08775 | 0.06772    |
| log(Seniors)  | 0.50315  | 0.10756    |

Below is a table of the VIF-values for each predictor before excluding a variable.

| Predictor     | VIF-values |
|---------------|------------|
| log(Fuel)     | 1.401090   |
| log(GRP)      | 1.304027   |
| log(Income)   | 1.982527   |
| log(Vehicles) | 9.090458   |
| log(Builton)  | 11.120575  |
| log(Higheds)  | 2.899656   |
| log(Seniors)  | 2.710582   |

By observing the VIF-values, we can see that `log(Vehicles)` and `log(Builton)` have higher VIF-values, which indicates multicollinearity. This can make coefficient estimates unstable and harder to interpret. Therefore, it may be advisable to exclude one of these predictors.

After excluding `log(Builton)` from the model, the estimates are:

| Predictor     | Estimate | Std. Error |
|---------------|----------|------------|
| Intercept     | -7.25934 | 1.42744    |
| log(Fuel)     | 0.21710  | 0.01971    |
| log(GRP)      | 0.07498  | 0.04190    |
| log(Income)   | 0.17419  | 0.24138    |
| log(Vehicles) | 0.75389  | 0.07182    |
| log(Higheds)  | -0.12972 | 0.06589    |
| log(Seniors)  | 0.55470  | 0.10618    |

| Predictor     | VIF-values |
|---------------|------------|
| log(Fuel)     | 1.399104   |
| log(GRP)      | 1.291286   |
| log(Income)   | 1.959979   |
| log(Vehicles) | 2.619874   |
| log(Higheds)  | 2.700781   |
| log(Seniors)  | 2.599338   |

The second predictor that had a high VIF-value was log(Vehicles) with a VIF-value of 9.090458 with standard error 0.13271. After excluding log(Builton), the VIF-value for log(Vehicles) dropped to 2.619874 with a standard error of 0.07182, which is a significant improvement and indicates that we have reduced multicollinearity in our model. Remaining predictors have VIF-values that are below the threshold. This means that we can keep them in the model without worrying about multicollinearity.

## 2(d) Significance tests and explained variability

We work with Model 2(c) (model_six) which contains the six log-transformed x-variables log(Fuel), log(GRP), log(Income), log(Vehicles), log(Higheds), log(Seniors). We assume the standard linear regression assumptions: independent errors and normally distributed residuals. Note that Model 1(b) is nested inside Model 2(c), so a partial F-test is valid.

```{r}
#| label: q2d
#| message: false
#| warning: false

# t-test on log(Fuel) in Model 2(c)
summary(model_six)$coefficients["log(Fuel)", ]
model_six$df.residual

# Partial F-test: Model 1(b) vs Model 2(c)
anova(another_model, model_six)

# R^2 and adj R^2 for Model 2(c)
c(R2    = summary(model_six)$r.squared,
  adjR2 = summary(model_six)$adj.r.squared)
```

**t-test on log(Fuel) in Model 2(c).** We did a two-sided t-test on $\beta_{\log(\text{Fuel})}$ which gives $t = 0.2171/0.0197 \approx 11$ on $t_{283}$ with $p < 2\cdot 10^{-16}$, so we reject $H_0: \beta_{\log(\text{Fuel})} = 0$ and conclude that log(Fuel) still has a significant relationship with log(PM10) after adjusting for the other five variables.

**Partial F-test: Model 1(b) vs Model 2(c).** Model 1(b) is nested in Model 2(c), so we can test whether adding the five extra variables (log(GRP), log(Income), log(Vehicles), log(Higheds), log(Seniors)) improves the fit with a partial F-test.

- **Type of test:** partial F-test (ANOVA comparing two nested linear models).
- **Null hypothesis:** $H_0: \beta_{\log(\text{GRP})} = \beta_{\log(\text{Income})} = \beta_{\log(\text{Vehicles})} = \beta_{\log(\text{Higheds})} = \beta_{\log(\text{Seniors})} = 0$, i.e. none of the five added variables contributes beyond log(Fuel). Alternative: at least one is non-zero.
- **Test statistic:** $F \approx 112.38$ (from `anova(another_model, model_six)`).
- **Distribution under $H_0$:** $F(5,\,283)$ — numerator df = 5 (added predictors), denominator df $= n - p = 290 - 7 = 283$.
- **P-value:** $p \approx 4.3\cdot 10^{-65}$ (R prints it as `< 2.2e-16`).
- **Conclusion:** at $\alpha = 0.05$ we clearly reject $H_0$, so at least one of the five added variables has a significant linear relationship with log(PM10) when log(Fuel) is already in the model.

**Explained variability.** From the summary, $R^2 \approx 0.7914$ and $R^2_{\text{adj}} \approx 0.7870$. Hence Model 2(c) explains about **79.1 %** of the variability in log(PM10) between municipalities (≈ 78.7 % after adjusting for the number of predictors).

## 2(e)

```{r}
df$Part <- as.factor(df$Part)
table(df$Part)
model_eight<- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors) + df$Part, data = df)
summary(model_eight)


vif(model_eight) |> knitr::kable()                                                                        
```

Part 1: 140 obs

Part 2: 96 obs

Part 3: 54 obs

| Predictor     | Estimate | Std. Error |
|---------------|---------:|-----------:|
| (Intercept)   | -6.85697 |    1.45193 |
| log(Fuel)     |  0.22538 |    0.01955 |
| log(GRP)      |  0.08315 |    0.04133 |
| log(Income)   |  0.25515 |    0.24166 |
| log(Vehicles) |  0.55362 |    0.13839 |
| log(Builton)  |  0.12510 |    0.05027 |
| log(Higheds)  | -0.06857 |    0.07007 |
| log(Seniors)  |  0.45357 |    0.10837 |
| Part2         | -0.07334 |    0.03025 |
| Part3         | -0.10051 |    0.04533 |

## 2(f) Two-way interactions in the model with testing

In this last section of part 2 of th project, we would like to add all two-way interactions between Part and each of the six numerical variables to obtain a new model. For this added interactions, we are interested in testing at level $\alpha= 5$ %, different from zero.

```{r}

# Here we are creating our model for two-way interactions
interaction_model <- lm(log(PM10) ~ (log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors))*Part, data=df)
interaction_model_sum <- summary(interaction_model)

# Test interactions terms (reduced vs full)
anova(model_seven, interaction_model)
```

To test if any of the added interactions is significantly, at level $\alpha = 5$ %, we will perform a partial F-test to compare a model with and without interactions. Our null hypothesis is that the added interactions is zero, otherwise not equal to 0. The test-statistic is $6.2665$ with $8$ and $274$ degrees of freedom for the $F$-value. P-value for the test is $1.882\cdot 10^{-7}$. The conclusion is that we can reject $H_0$ at signifcance level $5$%. There are significant differences between interactions parameters.

## What does this mean?? its distribution when the null hypothesis is true (including the degrees of freedom)

## 2(g) Leverage Analysis

In this section, we analyze leverage values (hat values) to identify observations that have high influence on the model fit.

```{r}
pred <- predict(interaction_model)
hat <- hatvalues(interaction_model)
top3 <- order(hat, decreasing = TRUE)[1:3]

# Create a data frame for plotting
plot_data <- data.frame(
  pred = pred,
  hat = hat,
  is_top3 = FALSE
)
plot_data$is_top3[top3] <- TRUE

ggplot(plot_data, aes(x = pred, y = hat, color = is_top3)) +
  geom_point(size = ifelse(plot_data$is_top3, 3, 1)) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"), 
                     labels = c("FALSE" = "Regular points", "TRUE" = "Top 3 leverage")) +
  geom_hline(yintercept = 2*length(coef(interaction_model))/nrow(df),
             linetype = "dashed", color = "red") +
  geom_hline(yintercept = 3*length(coef(interaction_model))/nrow(df),
             linetype = "dashed", color = "blue") +
  labs(title = "Leverage vs Predicted Values",
       x = "Predicted values",
       y = "Leverage (hat values)",
       color = "Point type") +
  theme_minimal()
```

## 3(b) Cook's distance and DFBETAS

In this section, we are interested in calculating Cook's distance for the model that added all two-way interactions between Part and each of the six numerical variables (a.k.a Model 2(f)).

```{r}
# Model 2(f)
interaction_model

# Summary of model 2(f)
interaction_model_sum

pike_pred <- mutate(
  df, 
  yhat = predict(interaction_model),
  r = rstudent(interaction_model),
  v = hatvalues(interaction_model),
  D = cooks.distance(interaction_model)
)

# with 1/n and 2(p+1)/n horizontal lines:
# p+1 = 

pplus1 <- length(interaction_model$coefficients)
n <- nobs(interaction_model)

f1.pike <- pplus1
f2.pike <- interaction_model$df.residual
cook.limit.pike <- qf(0.5, f1.pike, f2.pike)
names(pike_pred)

# Find top 3 by Cook's D
top3_cook <- pike_pred |> slice_max(D, n = 3)

ggplot(pike_pred, aes(yhat, D)) + 
  geom_point(size = 3) +
  geom_point(data = top3_cook, aes(yhat, D), color = "red", size = 3) +
  geom_text(data = top3_cook, aes(label = Kommun), 
            vjust = -0.8, color = "red", size = 3) +
  geom_hline(yintercept = cook.limit.pike, color = "red") +
  geom_hline(yintercept = 4/n, linetype = 2, color = "red") +
  xlab("Fitted values") +
  ylab("D_i") +
  labs(title = "Pike: Cook's D",
       caption = "4/n (dashed), F_0.5, p+1, n-(p+1) (solid)")

# Here we compute DFBETAS for the beta estimation(s)
head(dfbetas(interaction_model))
```

From the figure above, we have plot Cook's distance, $D_i$, vs the linear predictor, $\widehat{Y}_i$. We have added the horizontal lines at $4/n$ (dashed) and $F_{0.5,p+1,n-(p+1)}$ (solid) for reference. We can observe that Oxelösund, Kiruna and Gällivare are the three municipalities with the highest Cook's distance. The three municipalities that had the highest leverage was Gällivare, Solna och Åre. This means that there are not the same municipalities with high leverage and high Cook's distance.

3c 
```{r}
pred <- predict(interaction_model)
r_star <- rstudent(interaction_model)

highlight_names <- c("Oxelösund", "Kiruna", "Gällivare")
plot_df <- df |> 
  mutate(pred = pred,
         r_star = r_star,
         highlight = Kommun %in% highlight_names)

highlight_df <- plot_df |> filter(highlight)

ggplot(plot_df, aes(x = pred, y = r_star)) +
  geom_point(data = filter(plot_df, !highlight), color = "black", size = 2) +
  geom_point(data = highlight_df, color = "red", size = 4) +
  geom_text(data = highlight_df, aes(label = Kommun), 
            vjust = -0.8, color = "red", size = 3) +
  geom_hline(yintercept = c(-3, 0, 3),
             linetype = c("dashed", "solid", "dashed"),
             color = "red") +
  xlab("Predicted values") +
  ylab("Standardized residuals") +
  labs(title = "Standardized residuals vs predicted values",
       subtitle = "Highlighted: Oxelösund, Kiruna, Gällivare") +
  theme_minimal()

df %>% filter(abs(r_star) > 3)

ggplot(df, aes(x = pred, y = sqrt(abs(r_star)))) +
  geom_point()
qqnorm(r_star)
qqline(r_star, col = "red")
```

**Top three Cook's D** (from `slice_max(D, n = 3)`): Oxelösund (D ≈ 0.47), Gällivare (D ≈ 0.26), Kiruna (D ≈ 0.13). All three are well above the $4/n$ reference line but below the $F_{0.5,\,p+1,\,n-(p+1)}$ line ($\approx 0.97$) — i.e. they are clearly flagged as influential but none is *individually* catastrophic.

**Comparison with leverage.** The top three leverage municipalities are Gällivare, Solna and Åre, so the leverage and Cook's D lists overlap only in Gällivare. Oxelösund and Kiruna have moderate leverage but large residuals, which is why their Cook's D dominates even though their hat values aren't extreme.

**DFBETAS for Oxelösund (highest Cook's D).** Looking at the full row (not `head()`, which only shows the first six municipalities), two coefficients exceed the $F_{0.5}$ line:

- **log(Fuel):Part2** — DFBETAS $\approx +1.49$
- **log(Vehicles):Part2** — DFBETAS $\approx -1.30$

Both are Part-2 (Svealand) interactions, which makes sense: Oxelösund is a small Svealand municipality hosting SSAB's blast furnace (Table 1), so its Fuel and Vehicle profiles are extreme relative to other Svealand municipalities. Removing it would shift the Svealand-specific slopes on log(Fuel) and log(Vehicles) by \~1.3–1.5 standard errors.

The follow-up plots of log(PM10) against log(Fuel) and log(Vehicles), with Oxelösund highlighted and coloured by Part, will make this visually obvious: Oxelösund sits far to the right of the Svealand cloud on both axes.

3d
```{r}
df_new <- filter(df, !Kommun %in% c("Oxelösund", "Kiruna", "Gällivare"))
model_new <- lm(log(PM10) ~ (log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors))*Part, data = df_new)
summary(model_new)

cook <- cooks.distance(model_new)
student2 <- rstudent(model_new)
pred2 <- predict(model_new)
r_star2 <- rstudent(model_new)

# Cook's distance plot for new model
ggplot(data.frame(pred2 = pred2, cook = cook), aes(x = pred2, y = cook)) +
  geom_point() +
  geom_hline(yintercept = 4/length(pred2), linetype = "dashed", color = "red") +
  labs(title = "Cook's distance for model excluding Oxelösund, Kiruna, Gällivare",
       x = "Predicted values", y = "Cook's distance")

# Residuals plot for new model
plot_df_new <- df_new |> 
  mutate(pred2 = pred2,
         r_star2 = r_star2)

ggplot(plot_df_new, aes(x = pred2, y = r_star2)) +
  geom_point(color = "black", size = 2) +
  geom_hline(yintercept = c(-3, 0, 3),
             linetype = c("dashed", "solid", "dashed"),
             color = "red") +
  xlab("Predicted values") +
  ylab("Standardized residuals") +
  labs(title = "Standardized residuals vs predicted values (new model)",
       subtitle = "Excluding Oxelösund, Kiruna, Gällivare") +
  theme_minimal()

# Check for outliers in new model
df_new %>% mutate(r_star2 = r_star2) %>% filter(abs(r_star2) > 3)
```



## 3(e) AIC and BIC
The goal of this section is to perform a stepwise selection using AIC and BIC criterion. 