---
title: "Project 1"
author: "Björn Ahl, Leon Krasniqi, Nora Granström"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

# Part 1. Particle emissions and fuel consumption

## 1(a) Choice of transformations

We compare residual plots for three candidate models to motivate the log of the response and to decide between `x = Fuel` and `x = log(Fuel)`.

```{r}
#| label: fig-1a-resid
#| fig-cap: "Residuals vs fitted for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

library(broom)
library(patchwork)

m_raw    <- lm(PM10      ~ Fuel,       data = df)
m_logy   <- lm(log(PM10) ~ Fuel,       data = df)
m_loglog <- lm(log(PM10) ~ log(Fuel),  data = df)

plot_resid <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = .6) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x, colour = "steelblue") +
    labs(title = title, x = "fitted", y = "residual")
}

plot_resid(m_raw,    "PM10 ~ Fuel") +
  plot_resid(m_logy,   "log(PM10) ~ Fuel") +
  plot_resid(m_loglog, "log(PM10) ~ log(Fuel)")
```

**Interpretation:**

- **Linear relationship.** In this case we have `PM10 = x₀ + x₁·Fuel`. From the graph we observe a stronger tendency toward the left, meaning observations cluster at lower fuel values while the spread of PM10 widens with Fuel. This suggests a purely linear model is not sufficient.
- **Logarithmic response.** `ln(PM10) = x₀ + x₁·Fuel` is equivalent to `PM10 = a·b^Fuel`. The residual plot is less skewed, but equal additive changes in Fuel are treated the same across the scale, which is unrealistic for municipalities of different sizes.
- **Log-log relationship.** `ln(PM10) = x₀ + x₁·ln(Fuel)` implies `PM10 = a·Fuel^{x₁}`. A relative change in Fuel implies a relative change in PM10, and the residuals are more evenly spread. This makes the log-log transformation the most plausible choice.

```{r}
#| label: fig-1a-qq
#| fig-cap: "QQ-plots of residuals for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

plot_qq <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(sample = .resid)) +
    stat_qq(alpha = .6) +
    stat_qq_line(colour = "steelblue") +
    labs(title = title, x = "theoretical quantiles", y = "sample quantiles")
}

plot_qq(m_raw,    "PM10 ~ Fuel") +
  plot_qq(m_logy,   "log(PM10) ~ Fuel") +
  plot_qq(m_loglog, "log(PM10) ~ log(Fuel)")
```

From the QQ-plots we can clearly see that the log–log transformation gives the best fit, as its residuals follow the reference line most closely and are therefore the most normally distributed of the three.


## 1(b) Confidence interval, prediction interval and $\beta$-estimates

In this part-question, we obtain the $\beta$-estimates and their 95% confidence intervals when `x = ln(Fuel)`.

```{r}

mymodel <- lm(PM10 ~ log(Fuel), data = df)
mymodel_summary <- summary(mymodel)
mymodel$coefficients
confint(mymodel)

another_model <- lm(log(PM10) ~ log(Fuel), data = df)

new_data <- data.frame(Fuel = seq(min(df$Fuel), max(df$Fuel), length.out = 100))

pred_int <- predict(another_model, newdata = new_data, interval = "prediction")
conf_int <- predict(another_model, newdata = new_data, interval = "confidence")

plot_df <- data.frame(
  Fuel = new_data$Fuel,
  fit = conf_int[, "fit"],
  ci_lower = conf_int[, "lwr"],
  ci_upper = conf_int[, "upr"],
  pi_lower = pred_int[, "lwr"],
  pi_upper = pred_int[, "upr"]
)

# Plot on log scale
p1 <- ggplot() +
  geom_point(data = df, aes(x = log(Fuel), y = log(PM10))) +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df, aes(x = log(Fuel), y = fit), color = "red") +
  labs(x = "ln(Fuel)", y = "ln(PM10)", title = "ln(PM10) vs ln(Fuel)") +
  theme_minimal()

# Transforming back to original scale
plot_df_e <- data.frame(
  Fuel = new_data$Fuel,
  fit = exp(conf_int[, "fit"]),
  ci_lower = exp(conf_int[, "lwr"]),
  ci_upper = exp(conf_int[, "upr"]),
  pi_lower = exp(pred_int[, "lwr"]),
  pi_upper = exp(pred_int[, "upr"])
)

p2 <- ggplot() +
  geom_point(data = df, aes(x = Fuel, y = PM10)) +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df_e, aes(x = Fuel, y = fit), color = "red") +
  labs(x = "Fuel", y = "PM10", title = "PM10 vs Fuel") +
  theme_minimal()

p1
p2

intercept <- cbind(
  expbeta = exp(another_model$coefficients["(Intercept)"]),
  exp(confint(another_model, parm = "(Intercept)"))
) |>
  round(digits = 4)

slope <- cbind(
  expbeta = exp(another_model$coefficients["log(Fuel)" ]),
  exp(confint(another_model, parm = "log(Fuel)"))
) |>
  round(digits = 4)
```

## 1(c) Effect of changing fuel consumption

Model 1(b) is the log–log model
$$\ln(\text{PM10}) = \beta_0 + \beta_1 \ln(\text{Fuel}) \;\Longleftrightarrow\; \text{PM10} = a \cdot \text{Fuel}^{\beta_1}, \quad a = e^{\beta_0}.$$

Here $\beta_1$ is an elasticity: if Fuel is multiplied by a factor $k$, PM10 is multiplied by $k^{\beta_1}$.

**Change rate for a 10 % decrease in Fuel.** Let $k = 0.9$, so the new Fuel is $0.9\,\text{Fuel}_{\text{old}}$. Then
$$\frac{\text{PM10}_{\text{new}}}{\text{PM10}_{\text{old}}} = \left(\frac{0.9\,\text{Fuel}_{\text{old}}}{\text{Fuel}_{\text{old}}}\right)^{\beta_1} = 0.9^{\beta_1},$$
giving the point estimate $\widehat{r} = 0.9^{\widehat{\beta}_1}$ and percent change $(0.9^{\widehat{\beta}_1}-1)\cdot 100\%$.

With the 95 % CI for $\beta_1$ given by $[L,\,U] = \widehat{\beta}_1 \pm t_{0.975,\,n-2}\cdot \text{SE}(\widehat{\beta}_1)$, the map $\beta_1 \mapsto 0.9^{\beta_1}$ is **decreasing** (since $0.9 < 1$), so
$$\text{CI}_{95\%}(r) = \bigl[\,0.9^{U},\;0.9^{L}\,\bigr].$$

**Fuel reduction required to halve PM10.** We solve $k^{\beta_1} = 0.5$ for $k$:
$$k = 0.5^{1/\beta_1}, \qquad \text{required reduction} = (1 - 0.5^{1/\beta_1})\cdot 100\%.$$

The map $\beta_1 \mapsto 0.5^{1/\beta_1}$ is **increasing** for $\beta_1 > 0$, so
$$\text{CI}_{95\%}(k) = \bigl[\,0.5^{1/L},\;0.5^{1/U}\,\bigr],$$
and the CI for the reduction is $\bigl[\,1 - 0.5^{1/U},\;1 - 0.5^{1/L}\,\bigr]\cdot 100\%.$

```{r}
#| label: q1c
#| message: false
#| warning: false

b1  <- coef(another_model)["log(Fuel)"]
ci  <- confint(another_model, parm = "log(Fuel)")
L   <- ci[1]; U <- ci[2]

# (i) 10 % decrease in Fuel
r_hat <- 0.9^b1
r_ci  <- c(0.9^U, 0.9^L)
pct_change <- (c(r_hat, r_ci) - 1) * 100

# (ii) fuel reduction to halve PM10
k_hat <- 0.5^(1/b1)
k_ci  <- c(0.5^(1/L), 0.5^(1/U))
reduction <- (1 - c(k_hat, k_ci[2], k_ci[1])) * 100

round(rbind(
  "PM10 change (%) from -10% Fuel" = pct_change,
  "Fuel reduction (%) to halve PM10" = reduction
), 2)
```

