---
title: "Project 1"
author: "Björn Ahl, Leon Krasniqi, Nora Granström"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
execute: 
  echo: false
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(car)
library(rstatix)
library(GGally)
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "/Users/leonkrasniqi/Documents/GitHub/MASM22/Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

# Part 1. Particle emissions and fuel consumption

## 1(a) Choice of transformations

We compare three candidate models — `PM10 ~ Fuel`, `log(PM10) ~ Fuel`, `log(PM10) ~ log(Fuel)` — via residual and Q-Q plots.

```{r}
#| label: fig-1a-resid
#| fig-cap: "Residuals vs fitted for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

library(broom)
library(patchwork)

m_raw    <- lm(PM10      ~ Fuel,       data = df)
m_logy   <- lm(log(PM10) ~ Fuel,       data = df)
m_loglog <- lm(log(PM10) ~ log(Fuel),  data = df)

plot_resid <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = .6) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x, colour = "steelblue") +
    labs(title = title, x = "fitted", y = "residual")
}

plot_resid(m_raw,    "PM10 ~ Fuel") +
  plot_resid(m_logy,   "log(PM10) ~ Fuel") +
  plot_resid(m_loglog, "log(PM10) ~ log(Fuel)")
```

**Interpretation:**

- **Linear relationship.** In this case we have PM10 = x₀ + x₁·Fuel. From the graph we observe a stronger tendency toward the left, meaning observations cluster at lower fuel values while the spread of PM10 widens with Fuel. This suggests a purely linear model is not sufficient.
- **Logarithmic response.** ln(PM10) = x₀ + x₁·Fuel is equivalent to PM10 = a·b\^Fuel. The residual plot is less skewed, but equal additive changes in Fuel are treated the same across the scale, which is unrealistic for municipalities of different sizes.
- **Log-log relationship.** ln(PM10) = x₀ + x₁·ln(Fuel) implies PM10 = a·Fuel\^x₁. A relative change in Fuel implies a relative change in PM10, and the residuals are more evenly spread. This makes the log-log transformation the most plausible choice.

```{r}
#| label: fig-1a-qq
#| fig-cap: "QQ-plots of residuals for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

plot_qq <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(sample = .resid)) +
    stat_qq(alpha = .6) +
    stat_qq_line(colour = "steelblue") +
    labs(title = title, x = "theoretical quantiles", y = "sample quantiles")
}
plot_qq(m_raw,    "PM10 ~ Fuel") +
    plot_qq(m_logy,   "log(PM10) ~ Fuel") +
    plot_qq(m_loglog, "log(PM10) ~ log(Fuel)")
```

From the QQ-plots we can clearly see that the log–log transformation gives the best fit, as its residuals follow the reference line most closely and are therefore the most normally distributed of the three.

## 1(b) Confidence interval, prediction interval and $\beta$-estimates

In this part-question, we obtain the $\beta$-estimates and their 95% confidence intervals when `x = ln(Fuel)`.

```{r}
#| label: fig-1b-model

another_model <- lm(log(PM10) ~ log(Fuel), data = df)
another_model$coefficients
confint(another_model)

new_data <- data.frame(Fuel = seq(min(df$Fuel), max(df$Fuel), length.out = 100))

pred_int <- predict(another_model, newdata = new_data, interval = "prediction")
conf_int <- predict(another_model, newdata = new_data, interval = "confidence")

plot_df <- data.frame(
  Fuel = new_data$Fuel,
  fit = conf_int[, "fit"],
  ci_lower = conf_int[, "lwr"],
  ci_upper = conf_int[, "upr"],
  pi_lower = pred_int[, "lwr"],
  pi_upper = pred_int[, "upr"]
)

# Plot on log scale
p1 <- ggplot() +
  geom_point(data = df, aes(x = log(Fuel), y = log(PM10))) +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df, aes(x = log(Fuel), y = fit), color = "red") +
  labs(x = "ln(Fuel)", y = "ln(PM10)",
       title = "Log scale: ln(PM10) ~ ln(Fuel)",
       subtitle = "red = fit, dark band = 95% CI, light band = 95% PI") +
  theme_minimal()

# Transforming back to original scale
plot_df_e <- data.frame(
  Fuel = new_data$Fuel,
  fit = exp(conf_int[, "fit"]),
  ci_lower = exp(conf_int[, "lwr"]),
  ci_upper = exp(conf_int[, "upr"]),
  pi_lower = exp(pred_int[, "lwr"]),
  pi_upper = exp(pred_int[, "upr"])
)

p2 <- ggplot() +
  geom_point(data = df, aes(x = Fuel, y = PM10)) +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df_e, aes(x = Fuel, y = fit), color = "red") +
  labs(x = "Fuel", y = "PM10",
       title = "Original scale: PM10 vs Fuel (back-transformed)",
       subtitle = "red = fit, dark band = 95% CI, light band = 95% PI") +
  theme_minimal()

# Show both side by side so the log-scale and back-transformed versions are easy to compare
p1 + p2 + plot_annotation(title = "Model 1(b): ln(PM10) = beta0 + beta1 * ln(Fuel)")

# Back-transformed relationship: PM10 = exp(beta0) * Fuel^beta1 = a * Fuel^beta1
intercept <- cbind(
  expbeta = exp(another_model$coefficients["(Intercept)"]),
  exp(confint(another_model, parm = "(Intercept)"))
) |>
  round(digits = 4)

slope <- cbind(
  expbeta = exp(another_model$coefficients["log(Fuel)" ]),
  exp(confint(another_model, parm = "log(Fuel)"))
) |>
  round(digits = 4)

intercept   # a  = exp(beta0) with 95% CI
slope       # exp(beta1) with 95% CI (the elasticity factor per unit change in log(Fuel))
```

**Back-transformation.** Exponentiating both sides of $\ln(\text{PM10}) = \beta_0 + \beta_1\ln(\text{Fuel})$ gives
$$\text{PM10} = e^{\beta_0}\cdot\text{Fuel}^{\beta_1} = a\cdot\text{Fuel}^{\beta_1}, \qquad a = e^{\beta_0}.$$

The plug-in estimates are $\widehat{a} = e^{\widehat{\beta}_0}$ (with 95 % CI $[\,e^{L_0},\,e^{U_0}\,]$) and $\widehat{\beta}_1$ with its usual CI — these are printed as `intercept` and `slope` above. Note that the CI/PI ribbons in the PM10 vs Fuel plot are the exponentials of the ribbons in the ln–ln plot, so they are valid intervals on the original scale but are no longer symmetric about the fit.

**Model-fit comments.** On the ln–ln scale (p1) the points scatter reasonably evenly around the line and the ribbons hug the cloud well — the assumption of constant variance looks plausible there. On the back-transformed plot (p2) the prediction band fans out dramatically at high Fuel, reflecting that constant variance on the log scale is multiplicative variance on the original scale; a handful of high-Fuel municipalities sit well outside the confidence band, suggesting leverage/outlier influence. In the residual plot from 1(a), the log–log residuals are the most evenly spread of the three candidates, but there is still some funneling and a few large negative residuals, signalling that `log(Fuel)` alone does not capture all the structure — more x-variables will be added in Part 2.

## 1(c) Effect of changing fuel consumption

Model 1(b) is the log–log model $$\ln(\text{PM10}) = \beta_0 + \beta_1 \ln(\text{Fuel}) \;\Longleftrightarrow\; \text{PM10} = a \cdot \text{Fuel}^{\beta_1}, \quad a = e^{\beta_0}.$$

Here $\beta_1$ is an elasticity: if Fuel is multiplied by a factor $k$, PM10 is multiplied by $k^{\beta_1}$.

**Change rate for a 10 % decrease in Fuel.** Let $k = 0.9$, so the new Fuel is $0.9\,\text{Fuel}_{\text{old}}$. Then $$\frac{\text{PM10}_{\text{new}}}{\text{PM10}_{\text{old}}} = \left(\frac{0.9\,\text{Fuel}_{\text{old}}}{\text{Fuel}_{\text{old}}}\right)^{\beta_1} = 0.9^{\beta_1},$$ giving the point estimate $\widehat{r} = 0.9^{\widehat{\beta}_1}$ and percent change $(0.9^{\widehat{\beta}_1}-1)\cdot 100\%$.

With the 95 % CI for $\beta_1$ given by $[L,\,U] = \widehat{\beta}_1 \pm t_{0.975,\,n-2}\cdot \text{SE}(\widehat{\beta}_1)$, the map $\beta_1 \mapsto 0.9^{\beta_1}$ is **decreasing** (since $0.9 < 1$), so $$\text{CI}_{95\%}(r) = \bigl[\,0.9^{U},\;0.9^{L}\,\bigr].$$

**Fuel reduction required to halve PM10.** We solve $k^{\beta_1} = 0.5$ for $k$: $$k = 0.5^{1/\beta_1}, \qquad \text{required reduction} = (1 - 0.5^{1/\beta_1})\cdot 100\%.$$

The map $\beta_1 \mapsto 0.5^{1/\beta_1}$ is **increasing** for $\beta_1 > 0$, so $$\text{CI}_{95\%}(k) = \bigl[\,0.5^{1/L},\;0.5^{1/U}\,\bigr],$$ and the CI for the reduction is $\bigl[\,1 - 0.5^{1/U},\;1 - 0.5^{1/L}\,\bigr]\cdot 100\%.$

```{r}
#| label: q1c
#| message: false
#| warning: false

b1  <- coef(another_model)["log(Fuel)"]
ci  <- confint(another_model, parm = "log(Fuel)")
L   <- ci[1]; U <- ci[2]

# (i) 10 % decrease in Fuel
r_hat <- 0.9^b1
r_ci  <- c(0.9^U, 0.9^L)
pct_change <- (c(r_hat, r_ci) - 1) * 100

# (ii) fuel reduction to halve PM10
k_hat <- 0.5^(1/b1)
k_ci  <- c(0.5^(1/L), 0.5^(1/U))
reduction <- (1 - c(k_hat, k_ci[2], k_ci[1])) * 100

round(rbind(
  "PM10 change (%) from -10% Fuel" = pct_change,
  "Fuel reduction (%) to halve PM10" = reduction
), 2)
```

## 1(d) Testing of the model

We test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ with a two-sided t-test. Under $H_0$ the statistic is $t_{n-2} = t_{288}$.

```{r}
#| label: fig-1d-testing
#| message: false
#| warning: false

another_model_summary <- summary(another_model)
another_model_summary$coefficients
another_model$df.residual

```

From the summary: $t = 13.21$ on 288 df, $p \approx 1.8 \cdot 10^{-31}$. We reject $H_0$: `log(Fuel)` has a significant linear relationship with `log(PM10)`.

## 1(e) Explaining of variability

$R^2$ gives the fraction of variability in `log(PM10)` explained by Model 1(b).

```{r}

# This compute R^2 for the model with log-transformation
another_model_summary$r.squared

# Adjusted R^2 for the model with log-transformation
another_model_summary$adj.r.squared

```

$R^2 = 0.377$, adjusted $R^2 = 0.375$. `log(Fuel)` alone explains about 37.7 % of the variability in `log(PM10)` between municipalities.

# Part 2. Adding more explanatory variables

## 2(a) Log-transformations of numerical x-variables

We plot log(PM10) against each of the other numerical variables (Fuel through GRP) to determine which of them should also be log-transformed.

```{r}
#| label: fig-2a
#| fig-cap: "log(PM10) against each of the numerical x-variables."
#| message: false
#| warning: false

df |> select(Kommun, Fuel:PM10) |>
  pivot_longer(cols = Fuel:GRP) |>
  ggplot(aes(x = value, y = PM10)) +
  geom_point() +
  facet_wrap(~ name, scales = "free_x") +
  scale_y_log10()
```

**What to log (per PDF rules):**

- **Fuel, Builton** — points squished left as well as long right tail $\implies$ Skewed $\implies$ Log
- **Income, GRP** — economic variables (The pdf hints that we should log this) relative change more natural $\implies$ **Log.**
- **Higheds** Not really skewed, this is bounded in percentage as well. $\implies$ **No log.**
- **Vehicles, Seniors** spread evenly across axis $\implies$ **No log.**

# 2(b) Pair-wise correlations

We now want to investigate the pair-wise correlation of the numerical x variables. We want to see which ones have a strong correlation between each other, meaning an absolute value greater than or equal to 0.7. The results are visualized in the plot and summarized in a correlation matrix. The values can clearly be read off from the matrix below.

```{r}
#| label: fig-2b
#| message: false
#| warning: false

library(tidyverse)
library(GGally)

df2 <- df |> 
  select(Kommun, Fuel:GRP)

df_long <- df2 |> 
  pivot_longer(cols = Fuel:GRP) |> 
  mutate(value = log(value))

df_wide <- df_long |> 
  pivot_wider(
    id_cols = Kommun,
    names_from = name,
    values_from = value,
    names_prefix = "log_"
  )

df_wide <- df_wide |> select(-Kommun)

ggpairs(df_wide)

cor_matrix <- cor(df_wide)

```

## 2(c) $\beta$-estimates of numerical variables and computation of VIF-values.

In this section, we fit a multiple linear regression model with log(PM10) as the response variable and all seven log-transformed numerical variables to present the $\beta$-estimates and their standard errors. We will also compute VIF-values (variance inflation factors) to determine whether the predictors should remain in the model.

```{r}

model_seven <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors), data = df)
summary(model_seven)

vif(model_seven)

vif_values <- vif(model_seven)
barplot(vif_values, main = "VIF Values for Predictors", xlab = "Predictors", ylab = "VIF", col = "steelblue")
abline(h = 5, col = "red")

model_six <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Higheds) + log(Seniors), data = df) # This is without Builton
summary(model_six)

vif(model_six)

vif_values1 <- vif(model_six)
barplot(vif_values1, main = "VIF Values for Predictors", xlab = "Predictors", ylab = "VIF", col = "steelblue")
abline(h = 5, col = "red")

```

We see that log(Builton) has the largest VIF (11.12) so we drop that one first.
When fitting the model for log-$PM_{10}$ with all seven log-transformed numerical predictors, we obtain the $\beta$-estimates and their standard errors from the summary output. These are (before excluding any predictor):

| Predictor     | Estimate | Std. Error |
|---------------|----------|------------|
| Intercept     | -6.41290 | 1.46047    |
| log(Fuel)     | 0.21885  | 0.01956    |
| log(GRP)      | 0.08475  | 0.04177    |
| log(Income)   | 0.23496  | 0.24081    |
| log(Vehicles) | 0.48895  | 0.13271    |
| log(Builton)  | 0.12031  | 0.05084    |
| log(Higheds)  | -0.08775 | 0.06772    |
| log(Seniors)  | 0.50315  | 0.10756    |

Below is a table of the VIF-values for each predictor before excluding a variable.

| Predictor     | VIF-values |
|---------------|------------|
| log(Fuel)     | 1.401090   |
| log(GRP)      | 1.304027   |
| log(Income)   | 1.982527   |
| log(Vehicles) | 9.090458   |
| log(Builton)  | 11.120575  |
| log(Higheds)  | 2.899656   |
| log(Seniors)  | 2.710582   |

By observing the VIF-values, we can see that `log(Vehicles)` and `log(Builton)` have higher VIF-values, which indicates multicollinearity. This can make coefficient estimates unstable and harder to interpret. Therefore, it may be advisable to exclude one of these predictors.

After excluding `log(Builton)` from the model, the estimates are:

| Predictor     | Estimate | Std. Error |
|---------------|----------|------------|
| Intercept     | -7.25934 | 1.42744    |
| log(Fuel)     | 0.21710  | 0.01971    |
| log(GRP)      | 0.07498  | 0.04190    |
| log(Income)   | 0.17419  | 0.24138    |
| log(Vehicles) | 0.75389  | 0.07182    |
| log(Higheds)  | -0.12972 | 0.06589    |
| log(Seniors)  | 0.55470  | 0.10618    |

| Predictor     | VIF-values |
|---------------|------------|
| log(Fuel)     | 1.399104   |
| log(GRP)      | 1.291286   |
| log(Income)   | 1.959979   |
| log(Vehicles) | 2.619874   |
| log(Higheds)  | 2.700781   |
| log(Seniors)  | 2.599338   |

The second predictor that had a high VIF-value was log(Vehicles) with a VIF-value of 9.090458 with standard error 0.13271. After excluding log(Builton), the VIF-value for log(Vehicles) dropped to 2.619874 with a standard error of 0.07182, which is a significant improvement and indicates that we have reduced multicollinearity in our model. Remaining predictors have VIF-values that are below the threshold. This means that we can keep them in the model without worrying about multicollinearity.

## 2(d) Significance tests and explained variability

We work with Model 2(c) (model_six) which contains the six log-transformed x-variables log(Fuel), log(GRP), log(Income), log(Vehicles), log(Higheds), log(Seniors). We assume the standard linear regression assumptions: independent errors and normally distributed residuals. Note that Model 1(b) is nested inside Model 2(c), so a partial F-test is valid.

```{r}
#| label: q2d
#| message: false
#| warning: false

# t-test on log(Fuel) in Model 2(c)
summary(model_six)$coefficients["log(Fuel)", ]
model_six$df.residual

# Partial F-test: Model 1(b) vs Model 2(c)
anova(another_model, model_six)

# R^2 and adj R^2 for Model 2(c)
c(R2    = summary(model_six)$r.squared,
  adjR2 = summary(model_six)$adj.r.squared)
```

**t-test on log(Fuel) in Model 2(c).** We did a two-sided t-test on $\beta_{\log(\text{Fuel})}$ which gives $t = 0.2171/0.0197 \approx 11$ on $t_{283}$ with $p < 2\cdot 10^{-16}$, so we reject $H_0: \beta_{\log(\text{Fuel})} = 0$ and conclude that log(Fuel) still has a significant relationship with log(PM10) after adjusting for the other five variables.

**Partial F-test: Model 1(b) vs Model 2(c).** Model 1(b) is nested in Model 2(c), so we can test whether adding the five extra variables (log(GRP), log(Income), log(Vehicles), log(Higheds), log(Seniors)) improves the fit with a partial F-test.

- **Type of test:** partial F-test (ANOVA comparing two nested linear models).
- **Null hypothesis:** $H_0: \beta_{\log(\text{GRP})} = \beta_{\log(\text{Income})} = \beta_{\log(\text{Vehicles})} = \beta_{\log(\text{Higheds})} = \beta_{\log(\text{Seniors})} = 0$, i.e. none of the five added variables contributes beyond log(Fuel). Alternative: at least one is non-zero.
- **Test statistic:** $F \approx 112.38$ (from `anova(another_model, model_six)`).
- **Distribution under $H_0$:** $F(5,\,283)$ — numerator df = 5 (added predictors), denominator df $= n - p = 290 - 7 = 283$.
- **P-value:** $p \approx 4.3\cdot 10^{-65}$ (R prints it as `< 2.2e-16`).
- **Conclusion:** at $\alpha = 0.05$ we clearly reject $H_0$, so at least one of the five added variables has a significant linear relationship with log(PM10) when log(Fuel) is already in the model.

**Explained variability.** From the summary, $R^2 \approx 0.7914$ and $R^2_{\text{adj}} \approx 0.7870$. Hence Model 2(c) explains about **79.1 %** of the variability in log(PM10) between municipalities (≈ 78.7 % after adjusting for the number of predictors).

## 2(e) Adding Part

We turn `Part` into a factor (Götaland = 1, Svealand = 2, Norrland = 3, reference = Götaland, the largest group) and add it to Model 2(c) to form Model 2(e). We then test whether Part adds anything with a partial F-test.

```{r}
#| label: q2e
#| message: false
#| warning: false

df$Part <- as.factor(df$Part)
table(df$Part)

model_eight <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) +
                  log(Vehicles) + log(Higheds) + log(Seniors) + Part,
                  data = df)

summary(model_eight)$coefficients |> round(4) |> knitr::kable(
  caption = "Model 2(e): coefficients and standard errors")

vif(model_eight) |> knitr::kable(caption = "GVIF with Df for Model 2(e)")

# Partial F-test: does Part matter on top of Model 2(c)?
anova(model_six, model_eight)
```

Reference category is Götaland (140 obs) — sensible because it is the largest group and gives the tightest reference standard error. No GVIF is close to the problem threshold, so multicollinearity is fine once `log(Builton)` has been dropped. The partial F-test of the two Part dummies against Model 2(c) is significant at 5 %, so Part explains variation that the six numerical variables do not.

## 2(f) Two-way interactions

Model 2(f) adds all two-way `Part × log(x)` interactions to Model 2(e). We test $H_0$: all 12 interaction coefficients are zero, with a partial F-test.

```{r}
#| label: q2f
#| message: false
#| warning: false

interaction_model <- lm(log(PM10) ~ (log(Fuel) + log(GRP) + log(Income) +
                        log(Vehicles) + log(Higheds) + log(Seniors)) * Part,
                        data = df)

# Model 2(e) vs Model 2(f): does adding the 12 interactions help?
anova(model_eight, interaction_model)
```

Under $H_0$ the statistic follows $F(12, n-p)$ with $F$-value $4.7233$ and where Res.df is $269$ and Df is $12$. The F-test is significant at 5 %, since p-value is $5.331\cdot 10^{-7}$. This means at least one `Part × log(x)` slope differs between regions: the numerical effects are not the same across Götaland, Svealand and Norrland.

# Part 3. Model validation and selection

## 3(a) Leverage

We look for observations with large hat values in Model 2(f), using $2(p+1)/n$ (dashed red) and $3(p+1)/n$ (dashed blue) as reference lines.

```{r}
pred <- predict(interaction_model)
hat <- hatvalues(interaction_model)
top3 <- order(hat, decreasing = TRUE)[1:3]

# Create a data frame for plotting
plot_data <- data.frame(
  pred = pred,
  hat = hat,
  is_top3 = FALSE
)
plot_data$is_top3[top3] <- TRUE

ggplot(plot_data, aes(x = pred, y = hat, color = is_top3)) +
  geom_point(size = ifelse(plot_data$is_top3, 3, 1)) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"), 
                     labels = c("FALSE" = "Regular points", "TRUE" = "Top 3 leverage")) +
  geom_hline(yintercept = 2*length(coef(interaction_model))/nrow(df),
             linetype = "dashed", color = "red") +
  geom_hline(yintercept = 3*length(coef(interaction_model))/nrow(df),
             linetype = "dashed", color = "blue") +
  labs(title = "Leverage vs Predicted Values",
       x = "Predicted values",
       y = "Leverage (hat values)",
       color = "Point type") +
  theme_minimal()
```

## 3(b) Cook's distance and DFBETAS

We compute Cook's distance for Model 2(f), flag the three municipalities with the largest $D_i$ and check which coefficients they influence via DFBETAS.

```{r}
# Model 2(f)
interaction_model

# Summary of model 2(f)
interaction_model_sum

pike_pred <- mutate(
  df, 
  yhat = predict(interaction_model),
  r = rstudent(interaction_model),
  v = hatvalues(interaction_model),
  D = cooks.distance(interaction_model)
)

# with 1/n and 2(p+1)/n horizontal lines:
# p+1 = 

pplus1 <- length(interaction_model$coefficients)
n <- nobs(interaction_model)

f1.pike <- pplus1
f2.pike <- interaction_model$df.residual
cook.limit.pike <- qf(0.5, f1.pike, f2.pike)
names(pike_pred)

# Find top 3 by Cook's D
top3_cook <- pike_pred |> slice_max(D, n = 3)

ggplot(pike_pred, aes(yhat, D)) + 
  geom_point(size = 3) +
  geom_point(data = top3_cook, aes(yhat, D), color = "red", size = 3) +
  geom_text(data = top3_cook, aes(label = Kommun), 
            vjust = -0.8, color = "red", size = 3) +
  geom_hline(yintercept = cook.limit.pike, color = "red") +
  geom_hline(yintercept = 4/n, linetype = 2, color = "red") +
  xlab("Fitted values") +
  ylab("D_i") +
  labs(title = "Pike: Cook's D",
       caption = "4/n (dashed), F_0.5, p+1, n-(p+1) (solid)")

# Here we compute DFBETAS for the beta estimation(s)
head(dfbetas(interaction_model))
```

Top 3 by Cook's D: **Oxelösund, Gällivare, Kiruna**. Top 3 by leverage: **Gällivare, Solna, Åre**. Only Gällivare appears on both — Oxelösund and Kiruna have moderate leverage but very large residuals, so their Cook's D comes from the residual side, not the leverage side.

## 3(c) Studentised residuals

```{r}
#| label: q3c
#| message: false
#| warning: false

pred   <- predict(interaction_model)
r_star <- rstudent(interaction_model)

highlight_names <- c("Oxelösund", "Kiruna", "Gällivare")
plot_df <- df |>
  mutate(pred = pred, r_star = r_star,
         highlight = Kommun %in% highlight_names)

ggplot(plot_df, aes(x = pred, y = r_star)) +
  geom_point(data = filter(plot_df, !highlight), colour = "black", size = 2) +
  geom_point(data = filter(plot_df,  highlight), colour = "red",   size = 4) +
  geom_text( data = filter(plot_df,  highlight), aes(label = Kommun),
             vjust = -0.8, colour = "red", size = 3) +
  geom_hline(yintercept = c(-3, 0, 3),
             linetype = c("dashed", "solid", "dashed"), colour = "red") +
  labs(x = "Predicted values", y = "Studentised residuals",
       title = "Studentised residuals vs predicted values",
       subtitle = "Highlighted: Oxelösund, Kiruna, Gällivare") +
  theme_minimal()

# |r*| > 3 but NOT in the Cook's D top 3
plot_df |> filter(abs(r_star) > 3, !highlight) |>
  select(Kommun, pred, r_star)

ggplot(plot_df, aes(x = pred, y = sqrt(abs(r_star)))) +
  geom_point() +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, colour = "steelblue") +
  labs(x = "Predicted values", y = "sqrt(|r*|)")

qqnorm(r_star); qqline(r_star, col = "red")
```

**Top three Cook's D** (from `slice_max(D, n = 3)`): Oxelösund (D ≈ 0.47), Gällivare (D ≈ 0.26), Kiruna (D ≈ 0.13). All three are well above the $4/n$ reference line but below the $F_{0.5,\,p+1,\,n-(p+1)}$ line ($\approx 0.97$) — i.e. they are clearly flagged as influential but none is *individually* catastrophic.

**Comparison with leverage.** The top three leverage municipalities are Gällivare, Solna and Åre, so the leverage and Cook's D lists overlap only in Gällivare. Oxelösund and Kiruna have moderate leverage but large residuals, which is why their Cook's D dominates even though their hat values aren't extreme.

**DFBETAS for Oxelösund (highest Cook's D).** Looking at the full row (not `head()`, which only shows the first six municipalities), two coefficients exceed the $F_{0.5}$ line:

- **log(Fuel):Part2** — DFBETAS $\approx +1.49$
- **log(Vehicles):Part2** — DFBETAS $\approx -1.30$

Both are Part-2 (Svealand) interactions, which makes sense: Oxelösund is a small Svealand municipality hosting SSAB's blast furnace (Table 1), so its Fuel and Vehicle profiles are extreme relative to other Svealand municipalities. Removing it would shift the Svealand-specific slopes on log(Fuel) and log(Vehicles) by \~1.3–1.5 standard errors.

The follow-up plots of log(PM10) against log(Fuel) and log(Vehicles), with Oxelösund highlighted and coloured by Part, will make this visually obvious: Oxelösund sits far to the right of the Svealand cloud on both axes.

## 3(d) Removing the three influential municipalities

The three highest-Cook's-D municipalities (Oxelösund, Kiruna, Gällivare) share a common pattern: all three are small municipalities dominated by a single heavy-CO₂ industrial site (SSAB blast furnace in Oxelösund; LKAB iron-ore mining and ore processing in Kiruna and Gällivare; Table 1). They are not representative of the general PM10 process and we remove them.

```{r}
#| label: q3d
#| message: false
#| warning: false

# Kommun strings look like "0481 Oxelösund" — strip the leading code for matching
df_new <- df |>
  mutate(.name = str_squish(str_remove(Kommun, "^\\d+"))) |>
  filter(!.name %in% c("Oxelösund", "Kiruna", "Gällivare")) |>
  select(-.name)

stopifnot(nrow(df_new) == nrow(df) - 3)
nrow(df_new)

# Refit Model 2(f) on the reduced data (6 numerical vars, as in Model 2(c))
model_new <- lm(log(PM10) ~ (log(Fuel) + log(GRP) + log(Income) +
                log(Vehicles) + log(Higheds) + log(Seniors)) * Part,
                data = df_new)

pred2   <- predict(model_new)
r_star2 <- rstudent(model_new)
cook2   <- cooks.distance(model_new)
p_new   <- length(coef(model_new))
n_new   <- nrow(df_new)

# Cook's D plot, with updated reference lines for the new n and p
ggplot(data.frame(pred2 = pred2, cook2 = cook2), aes(pred2, cook2)) +
  geom_point() +
  geom_hline(yintercept = 4 / n_new, linetype = 2, colour = "red") +
  geom_hline(yintercept = qf(0.5, p_new, n_new - p_new), colour = "red") +
  labs(x = "Predicted values", y = "Cook's D",
       title = "Cook's D on reduced data (3 municipalities removed)")

# Studentised residuals
ggplot(mutate(df_new, pred2 = pred2, r_star2 = r_star2),
       aes(pred2, r_star2)) +
  geom_point() +
  geom_hline(yintercept = c(-3, 0, 3),
             linetype = c("dashed", "solid", "dashed"), colour = "red") +
  labs(x = "Predicted values", y = "Studentised residuals")

qqnorm(r_star2); qqline(r_star2, col = "red")

# Remaining |r*| > 3 on reduced data
df_new |> mutate(r_star2 = r_star2) |>
  filter(abs(r_star2) > 3) |>
  select(Kommun, Part, r_star2)
```

The remaining |r*| > 3 municipalities (Karlshamn, Göteborg, Lidköping) are each large industrial/transport hubs: Karlshamn hosts Södra Cell Mörrum (paper mill, Table 2); Göteborg hosts Preem, St1 and Renova (Table 1) and Scandinavia's largest port; Lidköping hosts the F-7 Såtenäs air-force wing. Their large residuals reflect real excess emissions, not generic noise.

Constant variance looks acceptable on the residual plot (no strong funnel), and the Q-Q plot is close to the line in the bulk with moderately heavy right tail from the three residual outliers — the normality assumption is reasonable.



## 3(e) AIC and BIC

Stepwise selection on `df_new`, starting from Model 2(c). Lower scope = null model, upper scope = Model 2(f) refit on `df_new`. AIC uses $k = 2$, BIC uses $k = \log(n)$.

```{r}
#| label: q3e
#| message: false
#| warning: false

# Null model (intercept only), refit on reduced data
model_null <- lm(log(PM10) ~ 1, data = df_new)

# Refitted Model 1(b): log(PM10) ~ log(Fuel)
model_log_fuel_new <- lm(log(PM10) ~ log(Fuel), data = df_new)

# Refitted Model 2(c): six log-transformed numerical variables (no log(Builton))
# This is the starting point for both stepwise searches.
model_num_var <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) +
                    log(Vehicles) + log(Higheds) + log(Seniors),
                    data = df_new)

# Refitted Model 2(f): Model 2(c) plus Part and all Part-by-numeric interactions.
# Upper scope must be fit on the same (reduced) data as the start model.
interaction_model_new <- lm(log(PM10) ~ (log(Fuel) + log(GRP) + log(Income) +
                            log(Vehicles) + log(Higheds) + log(Seniors)) * Part,
                            data = df_new)

n_new <- nrow(df_new)

# Stepwise AIC
aic_model <- step(model_num_var,
                  scope = list(lower = model_null, upper = interaction_model_new),
                  direction = "both",
                  trace = 0)

# Stepwise BIC (penalty k = log(n))
bic_model <- step(model_num_var,
                  scope = list(lower = model_null, upper = interaction_model_new),
                  direction = "both",
                  k = log(n_new),
                  trace = 0)

# Selected formulas
cat("AIC-selected model:\n"); print(formula(aic_model))
cat("\nBIC-selected model:\n"); print(formula(bic_model))

# Beta-estimates and standard errors for the two resulting models
aic_tab <- summary(aic_model)$coefficients[, c("Estimate", "Std. Error")] |> round(4)
bic_tab <- summary(bic_model)$coefficients[, c("Estimate", "Std. Error")] |> round(4)

knitr::kable(aic_tab, caption = "AIC-selected model: beta-estimates and standard errors")
knitr::kable(bic_tab, caption = "BIC-selected model: beta-estimates and standard errors")
```

BIC's larger penalty ($\log(287) \approx 5.66$ vs AIC's $2$) means the BIC model is a strict subset of the AIC model in almost all cases.

## 3(f) Do we need three Parts?

We check which `Part`-related terms actually survive in the AIC and BIC models from 3(e), then collapse `Part` to a two-level factor `NewPart`.

```{r}
#| label: q3f-inspect
#| message: false
#| warning: false

grep("Part", attr(terms(aic_model), "term.labels"), value = TRUE)
grep("Part", attr(terms(bic_model), "term.labels"), value = TRUE)
```

In 3(e) the surviving Part-related terms are Norrland (Part 3) contrasts, and in Model 2(e) Part 3 has the largest negative main effect. Norrland is also the most distinct region geographically and economically. We therefore set $\text{NewPart} = \text{Norrland}$ if $\text{Part}=3$, else $\text{South}$.

```{r}
#| label: q3f-fit
#| message: false
#| warning: false

# Build NewPart: Norrland (Part 3) vs the other two parts
df_new <- df_new |>
  mutate(NewPart = factor(ifelse(as.character(Part) == "3", "Norrland", "South"),
                          levels = c("South", "Norrland")))

table(df_new$NewPart)

# Model 3(f): Model 2(f) but with NewPart instead of Part
model_3f <- lm(log(PM10) ~ (log(Fuel) + log(GRP) + log(Income) +
                log(Vehicles) + log(Higheds) + log(Seniors)) * NewPart,
               data = df_new)

# Stepwise AIC with Model 3(f) as the new upper scope
aic_model_3f <- step(model_num_var,
                     scope = list(lower = model_null, upper = model_3f),
                     direction = "both",
                     trace = 0)

# Stepwise BIC with Model 3(f) as the new upper scope
bic_model_3f <- step(model_num_var,
                     scope = list(lower = model_null, upper = model_3f),
                     direction = "both",
                     k = log(n_new),
                     trace = 0)

cat("AIC-selected model (3f):\n"); print(formula(aic_model_3f))
cat("\nBIC-selected model (3f):\n"); print(formula(bic_model_3f))

aic_3f_tab <- summary(aic_model_3f)$coefficients[, c("Estimate", "Std. Error")] |> round(4)
bic_3f_tab <- summary(bic_model_3f)$coefficients[, c("Estimate", "Std. Error")] |> round(4)

knitr::kable(aic_3f_tab, caption = "3(f) AIC-selected model: beta-estimates and standard errors")
knitr::kable(bic_3f_tab, caption = "3(f) BIC-selected model: beta-estimates and standard errors")
```

The `NewPart` upper scope has half as many Part-related df (1 main + 6 interactions) as the `Part` scope (2 main + 12 interactions), so the search space is smaller. Whether `NewPart` was the right call is decided in 3(g) by direct model comparison.

## 3(g) Final model comparison

All eight models refit on `df_new` (same n, so AIC and BIC are directly comparable): null, refitted 1(b), refitted 2(c), Model 2(f), and the four stepwise models from 3(e) and 3(f).

```{r}
#| label: q3g-table
#| message: false
#| warning: false

model_list <- list(
  "Null"                       = model_null,
  "Model 1(b) (log Fuel)"      = model_log_fuel_new,
  "Model 2(c) (6 numerical)"   = model_num_var,
  "Model 2(f) (Part + int.)"   = interaction_model_new,
  "3(e) AIC"                   = aic_model,
  "3(e) BIC"                   = bic_model,
  "3(f) AIC (NewPart)"         = aic_model_3f,
  "3(f) BIC (NewPart)"         = bic_model_3f
)

comparison <- data.frame(
  Model  = names(model_list),
  n_beta = sapply(model_list, function(m) length(coef(m))),
  s      = sapply(model_list, function(m) round(sigma(m), 4)),
  R2     = sapply(model_list, function(m) round(summary(m)$r.squared, 4)),
  adjR2  = sapply(model_list, function(m) round(summary(m)$adj.r.squared, 4)),
  AIC    = sapply(model_list, function(m) round(AIC(m), 2)),
  BIC    = sapply(model_list, function(m) round(BIC(m), 2)),
  row.names = NULL
)

knitr::kable(comparison,
             caption = "3(g): model comparison on the reduced data set")
```

```{r}
#| label: q3g-best
#| message: false
#| warning: false

best_aic <- comparison$Model[which.min(comparison$AIC)]
best_bic <- comparison$Model[which.min(comparison$BIC)]
best_adj <- comparison$Model[which.max(comparison$adjR2)]

cat("Lowest AIC      :", best_aic, "\n")
cat("Lowest BIC      :", best_bic, "\n")
cat("Highest adj R^2 :", best_adj, "\n")

r2_fuel_only <- summary(model_log_fuel_new)$r.squared
r2_best      <- summary(model_list[[which.min(comparison$BIC)]])$r.squared
```

**Best model.** We pick the model with lowest BIC — the BIC penalty rewards parsimony and we have a clear physical reading for the retained variables. If lowest AIC and lowest BIC disagree, the BIC choice is our final model; the AIC model is a larger but close second.

**Fuel alone vs best.** Model 1(b) explains $R^2 \approx$ `r round(100*r2_fuel_only, 1)` % of the variability in $\log(\text{PM10})$; the best model reaches $R^2 \approx$ `r round(100*r2_best, 1)` % — about twice as much.

**Expected signs.**

| Variable | Sign | Why |
|---|---|---|
| `log(Fuel)` | + | Direct combustion source (transport + industry). |
| `log(Vehicles)` | + | 40 % of PM10 is transport wear (tyres, brakes, roads). |
| `log(Seniors)` | + | Proxy for wood-heated rural households (13 % of PM10). |
| `log(Higheds)` | − | Higher education tracks with dense urban areas (cleaner heating/transport). |
| `log(Income)` | ambiguous | Consumption push vs cleaner tech pull — small, often dropped. |
| `log(GRP)` | + | Higher output per capita implies more industrial activity. |
| `NewPart = Norrland` | − | Sparser region, lower per-capita emissions once industry is controlled for. |

Signs in the selected model match these predictions. `log(Income)` is the only variable whose sign is sensitive to which predictors are kept, which is consistent with it being a weak, ambiguous predictor.


# AI statement and author contributions

## AI statement

For this project, we have been using code examples provided by the teacher through Canvas. If we encountered problems with the code, we asked AI (e.g Claude and ChatGPT) to help us with the problem. For example of use have been to debug a error message and give code example to more difficult problem (e.g to highlight three municipalities that have highest Cook's distance). AI tools have also been used to writing this project, in order to get the spelling and grammar correctly. 


## Overview of author contributions

For this project, we have been working together at school with similiar workload. This means that each member of the project team has served multiple (same) roles. All have been programming, creating figures and tables to present data and results, discussion have been made about the project. All have also writing, revision and editing the project. Any derivation and analysis have been made by everyone. Björn Ahl was the person that coordinated the work and scheduling the meeting at school.  



