---
title: "Project 1"
author: "Björn Ahl"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

## Exploratory summary

```{r}
head(df)
glimpse(df)
summary(df)
```

# Part 1. Particle emissions and fuel consumption

## 1(a) Choice of transformations

We compare residual plots for three candidate models to motivate the log of the response and to decide between `x = Fuel` and `x = log(Fuel)`.

```{r}
#| label: fig-1a-resid
#| fig-cap: "Residuals vs fitted for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

library(broom)
library(patchwork)

m_raw    <- lm(PM10      ~ Fuel,       data = df)
m_logy   <- lm(log(PM10) ~ Fuel,       data = df)
m_loglog <- lm(log(PM10) ~ log(Fuel),  data = df)

plot_resid <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = .6) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x, colour = "steelblue") +
    labs(title = title, x = "fitted", y = "residual")
}

plot_resid(m_raw,    "PM10 ~ Fuel") +
  plot_resid(m_logy,   "log(PM10) ~ Fuel") +
  plot_resid(m_loglog, "log(PM10) ~ log(Fuel)")
```

**Interpretation.** `PM10 ~ Fuel` shows a clear funnel (variance grows with the fit) — classic heteroscedasticity pointing to a log of the response. `log(PM10) ~ Fuel` removes the funnel but leaves curvature in the loess line because `Fuel` is right-skewed. `log(PM10) ~ log(Fuel)` flattens both the trend and the spread, so we take **x = log(Fuel)**.