---
title: "Project 1"
author: "Björn Ahl"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

## Exploratory summary

```{r}
head(df)
glimpse(df)
summary(df)
```

# Part 1. Particle emissions and fuel consumption

## 1(a) Choice of transformations

We compare residual plots for three candidate models to motivate the log of the response and to decide between `x = Fuel` and `x = log(Fuel)`.

```{r}
#| label: fig-1a-resid
#| fig-cap: "Residuals vs fitted for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

library(broom)
library(patchwork)

m_raw    <- lm(PM10      ~ Fuel,       data = df)
m_logy   <- lm(log(PM10) ~ Fuel,       data = df)
m_loglog <- lm(log(PM10) ~ log(Fuel),  data = df)

plot_resid <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = .6) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x, colour = "steelblue") +
    labs(title = title, x = "fitted", y = "residual")
}

plot_resid(m_raw,    "PM10 ~ Fuel") +
  plot_resid(m_logy,   "log(PM10) ~ Fuel") +
  plot_resid(m_loglog, "log(PM10) ~ log(Fuel)")
```

**Interpretation.** `PM10 ~ Fuel` shows a clear funnel (variance grows with the fit) — classic heteroscedasticity pointing to a log of the response. `log(PM10) ~ Fuel` removes the funnel but leaves curvature in the loess line because `Fuel` is right-skewed. `log(PM10) ~ log(Fuel)` flattens both the trend and the spread, so we take **x = log(Fuel)**.




## 1(b) Confidence interval, predicition interval and $\beta$-estimates

In this part-question, we are interested to obtain the $\beta$-estimates and their 95 % confidence intervals when x=ln(Fuel). 


```{r}

mymodel <- lm(PM10 ~ log(Fuel), data=df)
mymodel_summary <- summary(mymodel)
mymodel$coefficients
confint(mymodel)

another_model <- lm(log(PM10) ~ log(Fuel), data=df)

# This need for smoother curves
new_data <- data.frame(Fuel = seq(min(df$Fuel), max(df$Fuel), length.out = 100))

pred_int = predict(another_model, newdata = new_data, interval = "prediction")

conf_int <- predict(another_model, newdata = new_data, interval = "confidence")

plot_df <- data.frame(
  Fuel = new_data$Fuel,
  fit = conf_int[, "fit"],
  ci_lower = conf_int[, "lwr"],
  ci_upper = conf_int[, "upr"],
  pi_lower = pred_int[, "lwr"],
  pi_upper = pred_int[, "upr"]
)

# Plot
ggplot() +
  geom_point(data = df, aes(x = log(Fuel), y = log(PM10))) +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df, aes(x = log(Fuel), y = fit), color = "red") +
  labs(x = "ln(Fuel)", y = "ln(PM10)", title = "ln(PM10) vs ln(Fuel)") +
  theme_minimal()

# Transforming the data from log-scale
e_conf_int <- exp(predict(another_model, newdata = new_data, interval = "confidence"))

e_pred_int <- exp(predict(another_model, newdata = new_data, interval = "prediction"))

intercept <- cbind(expbeta = exp(another_model$coefficients["(Intercept)"]),
      exp(confint(another_model, parm = "(Intercept)"))) |>
  round(digits = 4)

slope <- cbind(expbeta = exp(another_model$coefficients["log(Fuel)"]),
      exp(confint(another_model, parm = "log(Fuel)"))) |>
  round(digits = 4)

plot_df_e <- data.frame(
  Fuel = new_data$Fuel,
  fit = e_conf_int[, "fit"],
  ci_lower = e_conf_int[, "lwr"],
  ci_upper = e_conf_int[, "upr"],
  pi_lower = e_pred_int[, "lwr"],
  pi_upper = e_pred_int[, "upr"]
)

ggplot() +
  geom_point(data = df, aes(x = Fuel, y = PM10)) +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df_e, aes(x = Fuel, y = fit), color = "red") +
  labs(x = "Fuel", y = "PM10", title = "PM10 vs Fuel") +
  theme_minimal()

```

