---
title: "Project 1"
author: "Björn Ahl, Leon Krasniqi, Nora Granström"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(car)
library(rstatix)
library(GGally)
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

# Part 1. Particle emissions and fuel consumption

## 1(a) Choice of transformations

We compare residual plots for three candidate models to motivate the log of the response and to decide between `x = Fuel` and `x = log(Fuel)`.

```{r}
#| label: fig-1a-resid
#| fig-cap: "Residuals vs fitted for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

library(broom)
library(patchwork)

m_raw    <- lm(PM10      ~ Fuel,       data = df)
m_logy   <- lm(log(PM10) ~ Fuel,       data = df)
m_loglog <- lm(log(PM10) ~ log(Fuel),  data = df)

plot_resid <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = .6) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x, colour = "steelblue") +
    labs(title = title, x = "fitted", y = "residual")
}

plot_resid(m_raw,    "PM10 ~ Fuel") +
  plot_resid(m_logy,   "log(PM10) ~ Fuel") +
  plot_resid(m_loglog, "log(PM10) ~ log(Fuel)")
```

**Interpretation:**

- **Linear relationship.** In this case we have `PM10 = x₀ + x₁·Fuel`. From the graph we observe a stronger tendency toward the left, meaning observations cluster at lower fuel values while the spread of PM10 widens with Fuel. This suggests a purely linear model is not sufficient.
- **Logarithmic response.** `ln(PM10) = x₀ + x₁·Fuel` is equivalent to `PM10 = a·b^Fuel`. The residual plot is less skewed, but equal additive changes in Fuel are treated the same across the scale, which is unrealistic for municipalities of different sizes.
- **Log-log relationship.** `ln(PM10) = x₀ + x₁·ln(Fuel)` implies `PM10 = a·Fuel^{x₁}`. A relative change in Fuel implies a relative change in PM10, and the residuals are more evenly spread. This makes the log-log transformation the most plausible choice.

```{r}
#| label: fig-1a-qq
#| fig-cap: "QQ-plots of residuals for PM10~Fuel, log(PM10)~Fuel, and log(PM10)~log(Fuel)."
#| message: false
#| warning: false

plot_qq <- function(mod, title) {
  augment(mod) |>
    ggplot(aes(sample = .resid)) +
    stat_qq(alpha = .6) +
    stat_qq_line(colour = "steelblue") +
    labs(title = title, x = "theoretical quantiles", y = "sample quantiles")
}

plot_qq(m_raw,    "PM10 ~ Fuel") +
  plot_qq(m_logy,   "log(PM10) ~ Fuel") +
  plot_qq(m_loglog, "log(PM10) ~ log(Fuel)")
```

From the QQ-plots we can clearly see that the log–log transformation gives the best fit, as its residuals follow the reference line most closely and are therefore the most normally distributed of the three.

## 1(b) Confidence interval, prediction interval and $\beta$-estimates

In this part-question, we obtain the $\beta$-estimates and their 95% confidence intervals when `x = ln(Fuel)`.

```{r}
#| label: fig-1b-model

mymodel <- lm(PM10 ~ log(Fuel), data = df)
mymodel_summary <- summary(mymodel)
mymodel$coefficients
confint(mymodel)

another_model <- lm(log(PM10) ~ log(Fuel), data = df)

new_data <- data.frame(Fuel = seq(min(df$Fuel), max(df$Fuel), length.out = 100))

pred_int <- predict(another_model, newdata = new_data, interval = "prediction")
conf_int <- predict(another_model, newdata = new_data, interval = "confidence")

plot_df <- data.frame(
  Fuel = new_data$Fuel,
  fit = conf_int[, "fit"],
  ci_lower = conf_int[, "lwr"],
  ci_upper = conf_int[, "upr"],
  pi_lower = pred_int[, "lwr"],
  pi_upper = pred_int[, "upr"]
)

# Plot on log scale
p1 <- ggplot() +
  geom_point(data = df, aes(x = log(Fuel), y = log(PM10))) +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df, aes(x = log(Fuel), ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df, aes(x = log(Fuel), y = fit), color = "red") +
  labs(x = "ln(Fuel)", y = "ln(PM10)", title = "ln(PM10) vs ln(Fuel)") +
  theme_minimal()

# Transforming back to original scale
plot_df_e <- data.frame(
  Fuel = new_data$Fuel,
  fit = exp(conf_int[, "fit"]),
  ci_lower = exp(conf_int[, "lwr"]),
  ci_upper = exp(conf_int[, "upr"]),
  pi_lower = exp(pred_int[, "lwr"]),
  pi_upper = exp(pred_int[, "upr"])
)

p2 <- ggplot() +
  geom_point(data = df, aes(x = Fuel, y = PM10)) +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = pi_lower, ymax = pi_upper),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(data = plot_df_e, aes(x = Fuel, ymin = ci_lower, ymax = ci_upper),
              alpha = 0.4, fill = "blue") +
  geom_line(data = plot_df_e, aes(x = Fuel, y = fit), color = "red") +
  labs(x = "Fuel", y = "PM10", title = "PM10 vs Fuel") +
  theme_minimal()

p1
p2

intercept <- cbind(
  expbeta = exp(another_model$coefficients["(Intercept)"]),
  exp(confint(another_model, parm = "(Intercept)"))
) |>
  round(digits = 4)

slope <- cbind(
  expbeta = exp(another_model$coefficients["log(Fuel)" ]),
  exp(confint(another_model, parm = "log(Fuel)"))
) |>
  round(digits = 4)
```

## 1(c) Effect of changing fuel consumption

Model 1(b) is the log–log model $$\ln(\text{PM10}) = \beta_0 + \beta_1 \ln(\text{Fuel}) \;\Longleftrightarrow\; \text{PM10} = a \cdot \text{Fuel}^{\beta_1}, \quad a = e^{\beta_0}.$$

Here $\beta_1$ is an elasticity: if Fuel is multiplied by a factor $k$, PM10 is multiplied by $k^{\beta_1}$.

**Change rate for a 10 % decrease in Fuel.** Let $k = 0.9$, so the new Fuel is $0.9\,\text{Fuel}_{\text{old}}$. Then $$\frac{\text{PM10}_{\text{new}}}{\text{PM10}_{\text{old}}} = \left(\frac{0.9\,\text{Fuel}_{\text{old}}}{\text{Fuel}_{\text{old}}}\right)^{\beta_1} = 0.9^{\beta_1},$$ giving the point estimate $\widehat{r} = 0.9^{\widehat{\beta}_1}$ and percent change $(0.9^{\widehat{\beta}_1}-1)\cdot 100\%$.

With the 95 % CI for $\beta_1$ given by $[L,\,U] = \widehat{\beta}_1 \pm t_{0.975,\,n-2}\cdot \text{SE}(\widehat{\beta}_1)$, the map $\beta_1 \mapsto 0.9^{\beta_1}$ is **decreasing** (since $0.9 < 1$), so $$\text{CI}_{95\%}(r) = \bigl[\,0.9^{U},\;0.9^{L}\,\bigr].$$

**Fuel reduction required to halve PM10.** We solve $k^{\beta_1} = 0.5$ for $k$: $$k = 0.5^{1/\beta_1}, \qquad \text{required reduction} = (1 - 0.5^{1/\beta_1})\cdot 100\%.$$

The map $\beta_1 \mapsto 0.5^{1/\beta_1}$ is **increasing** for $\beta_1 > 0$, so $$\text{CI}_{95\%}(k) = \bigl[\,0.5^{1/L},\;0.5^{1/U}\,\bigr],$$ and the CI for the reduction is $\bigl[\,1 - 0.5^{1/U},\;1 - 0.5^{1/L}\,\bigr]\cdot 100\%.$

```{r}
#| label: q1c
#| message: false
#| warning: false

b1  <- coef(another_model)["log(Fuel)"]
ci  <- confint(another_model, parm = "log(Fuel)")
L   <- ci[1]; U <- ci[2]

# (i) 10 % decrease in Fuel
r_hat <- 0.9^b1
r_ci  <- c(0.9^U, 0.9^L)
pct_change <- (c(r_hat, r_ci) - 1) * 100

# (ii) fuel reduction to halve PM10
k_hat <- 0.5^(1/b1)
k_ci  <- c(0.5^(1/L), 0.5^(1/U))
reduction <- (1 - c(k_hat, k_ci[2], k_ci[1])) * 100

round(rbind(
  "PM10 change (%) from -10% Fuel" = pct_change,
  "Fuel reduction (%) to halve PM10" = reduction
), 2)
```

## 1(d) Testing of the model

In this part, we are interested to test if there is a significant relationship between `log(PM10)` and `log(Fuel)`, according to the linear regression model. We can do this by using a t-test on the null hypothesis $H_0: \beta_1 = 0$ against the alternative $H_1: \beta_1 \neq 0$.

```{r}
#| label: fig-1d-testing
#| message: false
#| warning: false

another_model_summary <- summary(another_model)
another_model_summary$coefficients
another_model$df.residual

```

The value of the test-statistic for the slope coefficient `log(Fuel)` is given by the t-value in the summary output. The p-value for the slope coefficient `log(Fuel)` (Pr(\>\|t\|)=1.828846e-31) is very small (less than 0.05), which indicates that we can reject the null hypothesis and conclude that there is a statistically significant relationship between `log(PM10)` and `log(Fuel)`.

## Part 2(a) Log-transformations of numerical x-variables

We plot log(PM10) against each of the other numerical variables (Fuel through GRP) to determine which of them should also be log-transformed.

```{r}
#| label: fig-2a
#| fig-cap: "log(PM10) against each of the numerical x-variables."
#| message: false
#| warning: false

df |> select(Kommun, Fuel:PM10) |>
  pivot_longer(cols = Fuel:GRP) |>
  ggplot(aes(x = value, y = PM10)) +
  geom_point() +
  facet_wrap(~ name, scales = "free_x") +
  scale_y_log10()
```

**What to log (per PDF rules):**

- **Fuel, Builton** — points squished left as well as long right tail $\implies$ Skewed $\implies$ Log
- **Income, GRP** — economic variables (The pdf hints that we should log this) relative change more natural $\implies$ **Log.**
- **Higheds** Not really skewed, this is bounded in percentage as well. $\implies$ **No log.**
- **Vehicles, Seniors** spread evenly across axis $\implies$ **No log.**

# 2(b) Pair-wise correlations

We now want to investigate the pair-wise correlation of the numerical x variables. We want to see which ones have a strong correlation between each other, meaning an absolute value greater than or equal to 0.7. The results are visualized in the plot and summarized in a correlation matrix.

```{r}
#| label: fig-2b
#| message: false
#| warning: false

library(tidyverse)
library(GGally)

df2 <- df |> 
  select(Kommun, Fuel:GRP)

df_long <- df2 |> 
  pivot_longer(cols = Fuel:GRP) |> 
  mutate(value = log(value))

df_wide <- df_long |> 
  pivot_wider(
    id_cols = Kommun,
    names_from = name,
    values_from = value,
    names_prefix = "log_"
  )

df_wide <- df_wide |> select(-Kommun)

ggpairs(df_wide)

cor_matrix <- cor(df_wide)

```

## 2(c) $\beta$-estimates of numerical variables and computation of VIF-values.

In this section, we fit a multiple linear regression model with log(PM10) as the response variable and all seven log-transformed numerical variables to present the $\beta$-estimates and their standard errors. We will also compute VIF-values (variance inflation factors) to determine whether the predictors should remain in the model.

```{r}

model_seven <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors), data = df)
summary(model_seven)

vif(model_seven)

vif_values <- vif(model_seven)
barplot(vif_values, main = "VIF Values for Predictors", xlab = "Predictors", ylab = "VIF", col = "steelblue")
abline(h = 5, col = "red")

model_six <- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Higheds) + log(Seniors), data = df)
summary(model_six)

vif(model_six)

vif_values1 <- vif(model_six)
barplot(vif_values1, main = "VIF Values for Predictors", xlab = "Predictors", ylab = "VIF", col = "steelblue")
abline(h = 5, col = "red")

```

When fitting the model for log-$PM_{10}$ with all seven log-transformed numerical predictors, we obtain the $\beta$-estimates and their standard errors from the summary output. These are (before excluding any predictor):

Below is a table of the VIF-values for each predictor before excluding a variable.

| Predictor     | Estimate | Std. Error |
|---------------|----------|------------|
| Intercept     | -6.41290 | 1.46047    |
| log(Fuel)     | 0.21885  | 0.01956    |
| log(GRP)      | 0.08475  | 0.04177    |
| log(Income)   | 0.23496  | 0.24081    |
| log(Vehicles) | 0.48895  | 0.13271    |
| log(Builton)  | 0.12031  | 0.05084    |
| log(Higheds)  | -0.08775 | 0.06772    |
| log(Seniors)  | 0.50315  | 0.10756    |

| Predictor     | VIF-values |
|---------------|------------|
| log(Fuel)     | 1.401090   |
| log(GRP)      | 1.304027   |
| log(Income)   | 1.982527   |
| log(Vehicles) | 9.090458   |
| log(Builton)  | 11.120575  |
| log(Higheds)  | 2.899656   |
| log(Seniors)  | 2.710582   |

By observing the VIF-values, we can see that `log(Vehicles)` and `log(Builton)` have higher VIF-values, which indicates multicollinearity. This can make coefficient estimates unstable and harder to interpret. Therefore, it may be advisable to exclude one of these predictors.

After excluding `log(Builton)` from the model, the updated estimates are:

| Predictor     | Estimate | Std. Error |
|---------------|----------|------------|
| Intercept     | -7.25934 | 1.42744    |
| log(Fuel)     | 0.21710  | 0.01971    |
| log(GRP)      | 0.07498  | 0.04190    |
| log(Income)   | 0.17419  | 0.24138    |
| log(Vehicles) | 0.75389  | 0.07182    |
| log(Higheds)  | -0.12972 | 0.06589    |
| log(Seniors)  | 0.55470  | 0.10618    |

| Predictor     | VIF-values |
|---------------|------------|
| log(Fuel)     | 1.399104   |
| log(GRP)      | 1.291286   |
| log(Income)   | 1.959979   |
| log(Vehicles) | 2.619874   |
| log(Higheds)  | 2.700781   |
| log(Seniors)  | 2.599338   |

The second predictor that had a high VIF-value was log(Vehicles) with a VIF-value of 9.090458 with standard error 0.13271. After excluding log(Builton), the VIF-value for log(Vehicles) dropped to 2.619874 with a standard error of 0.07182, which is a significant improvement and indicates that we have reduced multicollinearity in our model. Remaining predictors have VIF-values that are below the threshold, which suggest that we can keep them in the model without worrying about multicollinearity.

## 2(d) Significance tests and explained variability

We work with Model 2(c) (model_six) which contains the six log-transformed x-variables log(Fuel), log(GRP), log(Income), log(Vehicles), log(Higheds), log(Seniors). We assume the standard linear regression assumptions: independent errors and normally distributed residuals. Note that Model 1(b) is nested inside Model 2(c), so a partial F-test is valid.

```{r}
#| label: q2d
#| message: false
#| warning: false

# t-test on log(Fuel) in Model 2(c)
summary(model_six)$coefficients["log(Fuel)", ]
model_six$df.residual

# R^2 and adj R^2 for Model 2(c)
c(R2    = summary(model_six)$r.squared,
  adjR2 = summary(model_six)$adj.r.squared)
```

We did a two-sided t-test on $\beta_{\log(\text{Fuel})}$ which gives $t = 0.2171/0.0197 \approx 11$ on $t_{283}$ with $p < 2\cdot 10^{-16}$, so we reject $H_0: \beta_{\log(\text{Fuel})} = 0$ and conclude that log(Fuel) still has a significant relationship with log(PM10) after adjusting for the other five variables.

**Explained variability.** From the summary, $R^2 \approx 0.7914$ and $R^2_{\text{adj}} \approx 0.7870$. Hence Model 2(c) explains about **79.1 %** of the variability in log(PM10) between municipalities (≈ 78.7 % after adjusting for the number of predictors).

## 2(e)

```{r}
df$Part <- as.factor(df$Part)
table(df$Part)
model_eight<- lm(log(PM10) ~ log(Fuel) + log(GRP) + log(Income) + log(Vehicles) + log(Builton) + log(Higheds) + log(Seniors) + df$Part, data = df)
summary(model_eight)


vif(model_eight) |> knitr::kable()                                                                        
```

Part 1: 140 obs

Part 2: 96 obs

Part 3: 54 obs

```         
          Estimate Std. Error
```

(Intercept) -6.85697 1.45193 log(Fuel) 0.22538 0.01955 log(GRP) 0.08315 0.04133 log(Income) 0.25515 0.24166 log(Vehicles) 0.55362 0.13839 log(Builton) 0.12510 0.05027 log(Higheds) -0.06857 0.07007 - log(Seniors) 0.45357 0.10837 df$Part2      -0.07334    0.03025
df$Part3 -0.10051 0.04533