---
title: "Project 2"
author: "Björn Ahl, Shihan He, Orsolya Bosáková"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
execute: 
  echo: false
  warning: false
---

## Setup

```{r}
library(tidyverse)
library(readxl)
library(car)
library(rstatix)
library(GGally)
```

```{r}
xlsx_path <- if (file.exists("Project1data.xlsx")) "Project1data.xlsx" else "/Users/bjorn_ahl/Downloads/GitHub/MASM22/Project_1/Project1data.xlsx"
df <- as.data.frame(read_excel(xlsx_path))
```

## 1(a)

```{r}
cutoff <- quantile(df$PM10, probs = 2/3)

df <- df |>
  mutate(
    pm10_high = as.numeric(PM10 >= cutoff),
    pm10_cat  = factor(pm10_high, levels = c(0, 1), labels = c("Low", "High"))
  )
```

```{r}
ggplot(df, aes(x = Fuel, y = pm10_high, color = pm10_cat)) +
  geom_point()
```

```{r}
table(df$pm10_cat)
cutoff
```

The cut-off value (upper tertile of PM10) is 5.3. This means that if an observation has value $\geq 5.3$ it will classify as 1 = "high" else 0 = "low".

## 1(b)

```{r}
df$Part <- factor(df$Part, levels = c(1, 2, 3), labels = c("Götaland", "Svealand", "Norrland"))
df$Part <- relevel(df$Part, ref = "Götaland")
```

```{r}
tab <- table(Part = df$Part, Emission = df$pm10_cat)
knitr::kable(tab)
```

```{r}
tbl1b <- data.frame(
  Part      = c("Götaland", "Svealand", "Norrland"),
  Low       = c(100, 72, 18),
  High      = c(40,  24, 36)
) |>
  mutate(
    p       = High / (Low + High),
    odds    = p / (1 - p),
    log_odds = log(odds),
    odds_ratios      = odds / odds[Part == "Götaland"],
    log_odds_ratios  = log(odds_ratios)
  )

knitr::kable(tbl1b, digits = 4,
             col.names = c("Part", "Low", "High", "p", "Odds", "Log-odds", "odds_ratios", "Log-odds_ratios"),
             caption = "Table 1(b). Götaland = ref for odds_ratios")
```

Going from Götaland to Norrland gives an odds ratio of 5 $\implies$ high emissions are 5x more likely in Norrland versus Götaland

## 1(c)

```{r}
# We need family = binomial so that we get logistic. Else we have default is normal distribution which would give just a linear model.
model_1c <- glm(pm10_high ~ Part, data = df, family = binomial)

ci_beta <- confint(model_1c)
ci_exp  <- exp(ci_beta)

tbl1c <- data.frame(
  beta    = coef(model_1c),
  SE      = summary(model_1c)$coef[, 2],
  CI_low  = ci_beta[, 1],
  CI_high = ci_beta[, 2],
  exp_beta     = exp(coef(model_1c)),
  exp_CI_low   = ci_exp[, 1],
  exp_CI_high  = ci_exp[, 2]
)

knitr::kable(tbl1c, digits = 4,
             col.names = c("Term", "β", "SE", "95% CI low", "95% CI high",
                           "e^β", "e^β 95% CI low", "e^β 95% CI high"),
             caption = "Table 1(c): Model 1(c) coefficient estimates")
```
Here we see that the $e^{\beta}$ correspond to the oods_ratios in the previous task. The intercept for our reference region (Götaland) is 0.4. This matches up with our "regular" odds for Götaland.

The log odds can really just be read off table 1(b). The model gives us standard errors and confidence intervals via the predict function. 

```{r}
nd   <- data.frame(Part = factor(c("Götaland", "Svealand", "Norrland"), levels = levels(df$Part)))
pred <- predict(model_1c, newdata = nd, type = "link", se.fit = TRUE)

lo    <- pred$fit
se    <- pred$se.fit
ci_lo <- lo - 1.96 * se
ci_hi <- lo + 1.96 * se

tbl_pred <- data.frame(
  Part      = nd$Part,
  log_odds  = lo,  SE = se,
  CI_low    = ci_lo, CI_high = ci_hi,
  prob      = plogis(lo),
  prob_low  = plogis(ci_lo), prob_high = plogis(ci_hi)
)

knitr::kable(tbl_pred, digits = 4,
             col.names = c("Part", "Log-odds", "SE", "CI low", "CI high",
                           "Probability", "Prob CI low", "Prob CI high"),
             caption = "Table 1(c): Probabilities per region from model_1c")
``` 

Norrland has a much higher probability of high emissions than Götaland and Svealand which are  similar to each other. Norrland also has a wider CI due to fewer observations, 54 vs 96 and 140. The overlap between Götaland and Svealand in the CI's  may actually imply that there is no real difference between them (probability-wise).

\newline
\newline
We should use a Likelyhood ratio test. 
Let $H_0: \beta{_1}= \beta{_2} =0 $ . I.e  that all three regions have the same probability of high emissions. (Götaland is our reference, hence no need to compute (it is our $\beta_{0}$))
```{r}
model_null <- glm(pm10_high ~ 1, data = df, family = binomial)
anova(model_null, model_1c, test = "LRT")
```
The value of the test statistic is the drop in deviance $= 29.4$. $H_0$ follows a $\chi^2(2)$ distribution (with 2 df). . The p-value is $4 \times 10^{-7}$, so we reject $H_0$. There is statistically strong evidence that the probability of high emissions differs between Götaland, Svealand and Norrland.

## 1(d)

```{r}
ggplot(df, aes(x = Fuel, y = pm10_high)) +
  geom_point(aes(color = pm10_cat)) +
  geom_smooth(se = FALSE)+
  ylim(0,1)
```

From this we see that as fuel increases we get that the probability of high emission increase as well (as defined in our low/high metric). In the beginning the curve rises rapidly and then plateuas. This is also the reason which motivates why we log-transform. The log transform sort of normalises the distribution and clustering in the beginning making it more linear. This is showcased in plot below.


```{r}
ggplot(df, aes(x = log(Fuel), y = pm10_high)) +
  geom_point(aes(color = pm10_cat)) +
  geom_smooth(se = FALSE) +
  ylim(0, 1)
```


\newline
We get a more symmetric/S-shaped curve and hence we should use it as an explanatory variable.

```{r}
model_1d <- glm(pm10_high ~ log(Fuel), data = df, family = binomial)

ci_beta <- confint(model_1d)
ci_exp  <- exp(ci_beta)

tbl1d <- data.frame(
  beta        = coef(model_1d),
  SE          = summary(model_1d)$coef[, 2],
  CI_low      = ci_beta[, 1],
  CI_high     = ci_beta[, 2],
  exp_beta    = exp(coef(model_1d)),
  exp_CI_low  = ci_exp[, 1],
  exp_CI_high = ci_exp[, 2]
)

knitr::kable(tbl1d, digits = 4,
             col.names = c("β", "SE", "95% CI low", "95% CI high",
                           "e^β", "e^β CI low", "e^β CI high"),
             caption = "Table 1(d): Model.1(d) coefficient estimates")
```
To then get our estimated probability of a high emission rate and its CI to our plot we then just add method = "glm" which then calls predict.glm under the hood. It then builds Wald's CI's on log odds and then backtransform with the plogis() function to get the CI on the probability scale (under the hood).


```{r}
ggplot(df, aes(x = log(Fuel), y = pm10_high)) +
  geom_point(aes(color = pm10_cat)) +
  geom_smooth(method = "glm", method.args = list(family = binomial)) +
  ylim(0, 1)
```

\newline
Here we also have a likelyhood ratio test. The reason for this is that the sample size is not incredibly large, generally only Wald tests are good if the sample size is large. Likeleyhood ratio test is better because it comprates the fit of the two models via the difference in deviance. 

\newline 
Our $H_0: \beta{log(Fuel)} = 0$ (Fuel has no effect on prob of high emissions).

```{r}
anova(model_null, model_1d, test = "LRT")
```
This yields difference in deviance = 50.35 (which is our test statistic).
Since the critical value is $\chi{^2}_{1}$ $\approx$ 3.84. Since 50.35 > 3.84 we can statistically reject that our $H_0: \beta{log(Fuel)} = 0$.

\newline
Just by the fact we have an exponential relationship. If log(Fuel) increases by 1 we get that Fuel increases by $e^1 - 1 = 1.72$ $\implies$ an increase of 172%. 
\newline
When the log-fuel increases by 1? The log-odds of high emissions is given by $\beta  + \beta \cdot ln(Fuel)$. (This jut follows from the fact that it models the log-odds as a linear function of the predictor.)
$$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta \cdot \log(\text{Fuel})$$  

Hence we get that the change is if ln(fuel) increases by 1 we get the the odds are given by $e^{1.24} \approx 3.46$ (since $\beta = 1.24$  in our model (see table 1(d))), with 95% CI $(e^{0.858},\, e^{1.671}) \approx (2.36,\, 5.32)$.
\newline
By similar logic when the Fuel consumption increases by 10% we get that it increases by $ln(1.1) ≈ 0.095$ and hence the odds multiply by $e^{(\beta \cdot 0.095)} \approx e^{0.118} \approx 1.126$. The CI would then also increase by this same margin. I.e $e^{\text{CI}\text{low} \cdot 0.095}$ and $e^{\text{CI}\text{high} \cdot 0.095}$ Reading off from the table we then get our 95% CI as (1.085, 1.173)

# 1(e)

```{r}
#| fig-width: 9
#| fig-height: 3.4
n <- nrow(df); pp <- length(coef(model_1d))

tibble(
  h        = hatvalues(model_1d),
  log_Fuel = log(df$Fuel),
  eta_hat  = predict(model_1d),
  p_hat    = predict(model_1d, type = "response")
) |>
  pivot_longer(-h) |>
  mutate(name = factor(name, c("log_Fuel", "eta_hat", "p_hat"),
                       labels = c("log(Fuel)", "linear predictor x·β̂", "p-hat"))) |>
  ggplot(aes(value, h)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 1/n,    linetype = "dashed", colour = "steelblue") +
  geom_hline(yintercept = 2*pp/n, linetype = "dashed", colour = "firebrick") +
  facet_wrap(~ name, scales = "free_x") +
  expand_limits(y = 0) +
  labs(y = "Leverage h", x = NULL,
       caption = sprintf("blue: 1/n = %.4f   red: 2(p+1)/n = %.4f", 1/n, 2*pp/n))
```

We see that a point further out has more advantage and can tip the fit more. But if the datapoints are too far out and the model is already saying $\hat p \approx 0$ or $\approx 1$ with pretty much full confidence then the points stop mattering. The two 0-points at $\log(\text{Fuel}) \approx 5.25$ land on the right bump because of where they sit on the x-axis and not because they happen to be 0.

In general the flat tails mean the model is confident ($\hat p \approx 0$ or $1$), so the working weight $\hat p(1-\hat p) \to 0$ there and observations in those tails barely move the fit which is why leverage drops back toward zero at the extremes.

For the reason why these valleys sit where they are is because the S-curve (from 1(d)) is steepest at the inflection ($\hat{p} =0.5$) but leverage also depends on the squared distance from the weighted mean. This yields the actual leverage. This is thus maximised when the data points are far from the weighted mean and not so far out that the model is certain.
# 1(f)

```{r}
ll_null <- as.numeric(logLik(model_null))
mcf_adj <- function(m) 1 - (as.numeric(logLik(m)) - length(coef(m))) / ll_null

tibble(
  Model = c("1(c): Part", "1(d): log(Fuel)"),
  K     = c(length(coef(model_1c)), length(coef(model_1d))),
  AIC   = c(AIC(model_1c), AIC(model_1d)),
  BIC   = c(BIC(model_1c), BIC(model_1d)),
  `McFadden adj Rsq` = c(mcf_adj(model_1c), mcf_adj(model_1d))
) |> knitr::kable(digits = 4, caption = "Table 1(f): model comparison")
```

All three criteria pick Model.1(d). Lower AIC and BIC, higher adjusted pseudo-$R^2$. Fuel consumption explains substantially more of the variation in high-emission probability than the geography factor and it also does so with fewer params.
